\documentclass[12pt,a4paper]{letter}    
\newcommand{\gal}{{\sf GALAHAD}}
%\makelabels
\include{cseletter}
\begin{document}
%\vspace*{-1.0in}
\begin{letter}{Professor Michael Saunders \\
Dept. of Management Science and Engineering \\
Terman Building \\
Stanford University \\
Stanford, CA 94305-4062, USA}

\noindent
\opening{Dear Michael,}
   
Thank you very much for your letter of  February 28th concerning the paper
\begin{center}
 \gal, a library of thread-safe Fortran 90 
packages for large-scale nonlinear optimization
\end{center}
that I wrote with Dominique Orban and Philippe Toint, and that
you have been handling for {\em ACM TOMS}.

We have now looked through your review, and have prepared a
revision that aims to take into account all of your comments.

In response to the specific comments:

\begin{enumerate}
\item We understand your perspective, but we feel that a change of
name at this stage would confuse both us and our user community.
Certainly, although we view some of the routines as being of perhaps
less global use that others, they all have their own documentation,
and we believe they stand on their own. Of course, each module has
its own internal procedures, but one of the nice features of
modern Fortran is that these may safely be hidden from users. We hope
you understand our reluctance to change after so long!

\item Done.

\item Done.

\item Done, although we called this $m^Q(s)$.

\item Yes, this was poorly written. We have tried to clarify this.

\item Comment added. The reason is simply that for the QPBAND case, 
the linear solver chose to perform additional pivoting for stability,
and this led to significant fill in.

\item Phrase removed.

\item Done.

\item Done.

\item Explained. We simply mean that the variable is not involved in
any of the general constraints.

\item This is indeed the case. We allow certain options to change
between presolving one problem and a succeeding `structurally similar'' one.

\item No. If an unsymmetric change is made to the extended matrix, it
is more efficient to be able to solve with both $A$ and its transpose.

\item Yes, they are updated too. Comment added.
\end{enumerate}
Although we still plan to add SQP solvers, we are still working on this,
so it is unfortunately a bit premature to add details now.

As for the marked-up manuscript, we have adopted all of your suggestions
except (old manuscript pages):

\begin{description}
\item p3. We have not reversed the orders of 
{\tt QPB} and {\tt QPA} as we believe that
{\tt QPB} is generally superior, and we really intend that
people try the latter before the former!

\item p7, eq (2.2) trailing comma retained.

\item p7, eq (2.3) trailing full-stop changed to a comma.
\end{description}

We all feel that you have done a superb job in reading the
paper very carefully, and are extremely grateful for your help.

\closing{With best wishes,}

\end{letter} 

\end{document}
