\begin{description}

\itt{error} is a scalar variable of type default \integer, that holds the
stream number for error messages. Printing of error messages in
{\tt \packagename\_solve} and {\tt \packagename\_terminate} is suppressed if
{\tt error} $\leq 0$.
The default is {\tt error = 6}.

\ittf{out} is a scalar variable of type default \integer, that holds the
stream number for informational messages. Printing of informational messages in
{\tt \packagename\_solve} is suppressed if {\tt out} $< 0$.
The default is {\tt out = 6}.

\itt{print\_level} is a scalar variable of type default \integer, that is used
to control the amount of informational output which is required. No
informational output will occur if {\tt print\_level} $\leq 0$. If
{\tt print\_level} $= 1$, a single line of output will be produced for each
iteration of the process. If {\tt print\_level} $\geq 2$, this output will be
increased to provide significant detail of each iteration.
The default is {\tt print\_level = 0}.

\itt{maxit} is a scalar variable of type default \integer, that holds the
maximum number of iterations which will be allowed in {\tt \packagename\_solve}.
The default is {\tt maxit = 1000}.

\itt{start\_print} is a scalar variable of type default \integer, that specifies
the first iteration for which printing will occur in {\tt \packagename\_solve}.
If {\tt start\_print} is negative, printing will occur from the outset.
The default is {\tt start\_print = -1}.

\itt{stop\_print} is a scalar variable of type default \integer, that specifies
the last iteration for which printing will occur in  {\tt \packagename\_solve}.
If {\tt stop\_print} is negative, printing will occur once it has been
started by {\tt start\_print}.
The default is {\tt stop\_print = -1}.

\itt{infeas\_max} is a scalar variable of type default \integer, that specifies
the number of iterations for which the overall infeasibility
of the problem is not reduced by at least a factor {\tt reduce\_infeas
before} the problem is flagged as infeasible (see {\tt reduce\_infeas}).
The default is {\tt infeas\_max = 200}.
%The default is {\tt infeas\_max = 10}.

\itt{muzero\_fixed} is a scalar variable of type default \integer,
that specifies
the number of iterations before the initial barrier parameter
(see {\tt muzero}) may be altered.
The default is {\tt muzero\_fixed = 1}.

\ifthenelse{\equal{\package}{cqp}}{
\itt{restore\_problem} is a scalar variable of type default \integer, that
specifies how much of the input problem is to be restored on output.
Possible values are:
\begin{description}
\itt{0} nothing is restored.
\itt{1} the vector data $\bmw$, $\bmg$,
   $\bmc^{l}$, $\bmc^{u}$, $\bmx^{l}$, and $\bmx^{u}$
   will be restored to their input values.
\itt{2} the entire problem, that is the above vector data along with
the Hessian $\bmH$, will be restored.
\end{description}
The default is {\tt restore\_problem = 2}.
}{}

\itt{indicator\_type} is a scalar variable of type default \integer,
that specifies the type of indicator used to assess when a variable or
constraint is active.
Possible values are:

\begin{description}
\itt{1} a variable/constraint is active if and only if the distance
to its neaerest bound is no larger than {\tt indicator\-\_tol\_p} (see below).
\itt{2} a variable/constraint is active if and only if the distance
to its neaerest bound is no larger than {\tt indicator\-\_tol\_pd}
(see below) times the magnitude of its corresponding dual variable.
\itt{3} a variable/constraint is active if and only if the distance
to its neaerest bound is no larger than {\tt indicator\-\_tol\_tapia}
(see below) times the distance to the same bound at the previous iteration.
\end{description}
The default is {\tt indicator\_type = 3}.

\itt{arc} is a scalar variable of type default \integer,
that specifies the type of residual trajectory used to define the path
to the solution.
Possible values are:

\begin{description}
\itt{1} the residual trajectory proposed by Zhang will be used.
\itt{2} the residual trajectory proposed by Zhao and Sun will be used; note
this trajectory does not necessarily ensure convergence, so should be used
with caution.
\itt{3} a combination in which Zhang's trajectory is used until the
method determines that Zhou and Sun's will be better.
\itt{4} a mixed linear-quadratic variant of Zhang's proposal will be used.
\end{description}
The default is {\tt arc = 1}.

\itt{series\_order} is a scalar variable of type default \integer, that
specifies the order of (Puiseux or Taylor) series to approximate the
residual trajectory. The default is {\tt series\_order = 2}.

\itt{infinity} is a scalar variable of type default \realdp, that is used to
specify which constraint bounds are infinite.
Any bound larger than {\tt infinity} in modulus will be regarded as infinite.
The default is {\tt infinity =} $10^{19}$.

%\itt{stop\_abs\_p} and {\tt stop\_rel\_p}
%are scalar variables of type default \realdp, that hold the
%required absolute and relative accuracy for the primal infeasibility
%(see Section~\ref{galmethod}).
%The absolute value of each component of the primal infeasibility
%on exit is required to be smaller than the larger of {\tt stop\_abs\_p} and
%{\tt stop\_rel\_p} times a ``typical value'' for this component.
%The defaults are {\tt stop\_abs\_p =} {\tt stop\_rel\_p =} $u^{1/3}$,
%where $u$ is {\tt EPSILON(1.0)} ({\tt EPSILON(1.0D0)} in
%{\tt \fullpackagename\_double}).

\itt{stop\_abs\_d} and {\tt stop\_rel\_d}
are scalar variables of type default \realdp, that hold the
required absolute and relative accuracy for the dual infeasibility
(see Section~\ref{galmethod}).
The absolute value of each component of the dual infeasibility
on exit is required to be smaller than the larger of {\tt stop\_abs\_p} and
{\tt stop\_rel\_p} times a ``typical value'' for this component.
The defaults are {\tt stop\_abs\_d =} {\tt stop\_rel\_d =} $u^{1/3}$,
where $u$ is {\tt EPSILON(1.0)} ({\tt EPSILON(1.0D0)} in
{\tt \fullpackagename\_double}).

\itt{stop\_abs\_c} and {\tt stop\_rel\_c}
are scalar variables of type default \realdp, that hold the
required absolute and relative accuracy
for the violation of complementary slackness
(see Section~\ref{galmethod}).
The absolute value of each component of the complementary slackness
on exit is required to be smaller than the larger of {\tt stop\_abs\_p} and
{\tt stop\_rel\_p} times a ``typical value'' for this component.
The defaults are {\tt stop\_abs\_c =} {\tt stop\_rel\_c =} $u^{1/3}$,
where $u$ is {\tt EPSILON(1.0)} ({\tt EPSILON(1.0D0)} in
{\tt \fullpackagename\_double}).

\itt{perturb\_h} is a scalar variable of type default \realdp, that specifies
any perturbation that is to be added to the diagonal of $\bmH$.
The default is {\tt perturb\_h = 0.0}.

\itt{prfeas} is a scalar variable of type default \realdp, that aims to specify
the closest that any initial variable may be to infeasibility. Any variable
closer to infeasibility than {\tt prfeas} will be moved to {\tt prfeas} from
the offending bound. However, if a variable is range bounded, and its bounds
are closer than {\tt prfeas} apart, it will be moved to the mid-point of the
two bounds.
The default is {\tt prfeas = $10^4$}.

\itt{dufeas} is a scalar variable of type default \realdp, that aims to specify
the closest that any initial dual variable or Lagrange multiplier may be to
infeasibility. Any variable closer to infeasibility than {\tt prfeas} will be
moved to {\tt dufeas} from the offending bound. However, if a dual variable
is range bounded, and its bounds are closer than {\tt dufeas} apart, it will
be moved to the mid-point of the two bounds.
The default is {\tt dufeas = $10^4$}.

\itt{muzero}  is a scalar variable of type default \realdp, that holds the
initial value of the barrier parameter. If {\tt muzero} is
not positive, it will be reset automatically to an appropriate value.
The default is {\tt muzero = -1.0}.

\itt{tau}  is a scalar variable of type default \realdp, that holds the
weight attached to dual infeasibility compared to complementarity
when assessing step acceptance.
The default is {\tt tau = 1.0}.

\itt{gamma\_c}  is a scalar variable of type default \realdp, that holds the
smallest value that individual complementarity pairs are allowed
to be relative to the average as the iteration proceeds.
The default is {\tt gamma\_c = $10^{-5}$}.

\itt{gamma\_f}  is a scalar variable of type default \realdp, that holds the
smallest value the average complementarity is allowed
to be relative to the dual infeasibility as the iteration proceeds.
The default is {\tt gamma\_c = $10^{-5}$}.

\itt{reduce\_infeas}  is a scalar variable of type default
\realdp, that specifies the
least factor by which the overall infeasibility of the problem must be reduced,
over {\tt infeas\_max} consecutive iterations,
for it not be declared infeasible (see {\tt infeas\_max)}.
The default is {\tt reduce\_infeas = 0.99}.

\itt{obj\_unbounded}  is a scalar variable of type default
\realdp, that specifies smallest
value of the objective function that will be tolerated before the problem
is declared to be unbounded from below.
The default is {\tt obj\_u\-nbounded =} $-u^{-2}$,
where $u$ is {\tt EPSILON(1.0)} ({\tt EPSILON(1.0D0)} in
{\tt \fullpackagename\_double}).

\itt{potential\_unbounded}  is a scalar variable of type default
\realdp, that specifies smallest
value of the potential function divided by the number of one-sided variable and
constraint bounds that will be tolerated before the analytic center is
declared to be unbounded.
The default is {\tt potential\_u\-nbounded = -10.0}.

\itt{identical\_bounds\_tol}
is a scalar variable of type default \realdp.
Every pair of constraint bounds
$(c_{i}^{l}, c_{i}^{u})$ or $(x_{j}^{l}, x_{j}^{u})$
that is closer than {\tt identical\_bounds\_tol}
will be reset to the average of their values,
$\half (c_{i}^{l} + c_{i}^{u})$ or $\half (x_{j}^{l} + x_{j}^{u})$
respectively.
The default is {\tt identical\_bounds\_tol =} $u$,
where $u$ is {\tt EPSILON(1.0)} ({\tt EPSILON(1.0D0)} in
{\tt \fullpackagename\_double}).

\itt{indicator\_tol\_p}
is a scalar variable of type default \realdp that
provides the indicator tolerance associated with the test
{\tt indicator\_type = 1}.
The default is {\tt indicator\_tol\_p =} $u^{1/3}$,
where $u$ is {\tt EPSILON(1.0)} ({\tt EPSILON(1.0D0)} in
{\tt \fullpackagename\_double}).

\itt{indicator\_tol\_pd}
is a scalar variable of type default \realdp that
provides the indicator tolerance associated with the test
{\tt indicator\_type = 2}.
The default is {\tt indicator\_tol\_pd =} 1.0.

\itt{indicator\_tol\_tapia}
is a scalar variable of type default \realdp that
provides the indicator tolerance associated with the test
{\tt indicator\_type = 3}.
The default is {\tt indicator\_tol\-\_tapia =} 0.9.

\itt{cpu\_time\_limit} is a scalar variable of type default \realdp,
that is used to specify the maximum permitted CPU time. Any negative
value indicates no limit will be imposed. The default is
{\tt cpu\_time\_limit = - 1.0}.

\itt{clock\_time\_limit} is a scalar variable of type default \realdp,
that is used to specify the maximum permitted elapsed system clock time.
Any negative value indicates no limit will be imposed. The default is
{\tt clock\_time\_limit = - 1.0}.

%\itt{remove\_dependencies} is a scalar variable of type
%default \logical, that must be set \true\ if the algorithm
%is to attempt to remove any linearly dependent constraints before
%solving the problem, and \false\ otherwise.
%We recommend removing linearly dependencies.
%The default is {\tt remove\_dependencies = .TRUE.}.

\itt{treat\_zero\_bounds\_as\_general} is a scalar variable of type
default \logical.
If it is set to \false, variables which
are only bounded on one side, and whose bound is zero,
will be recognised as non-negativities/non-positivities rather than simply as
lower- or upper-bounded variables.
If it is set to \true, any variable bound
$x_{j}^{l}$ or $x_{j}^{u}$ which has the value 0.0 will be
treated as if it had a general value.
Setting {\tt treat\_zero\_bounds\_as\_general} to \true\ has the advantage
that if a sequence of problems are reordered, then bounds which are
``accidentally'' zero will be considered to have the same structure as
those which are nonzero. However, {\tt \fullpackagename} is
able to take special advantage of non-negativities/non-positivities, so
if a single problem, or if a sequence of problems whose
bound structure is known not to change, is/are to be solved,
it will pay to set the variable to \false.
The default is {\tt treat\_zero\_bounds\_as\_general = .FALSE.}.

\itt{just\_feasible} is a scalar variable of type default \logical, that
must be set \true\ if the algorithm should stop as soon as a feasible point
of the constraint set is found, and \false\ otherwise.
%We recommend using the analytic center.
The default is {\tt just\_feasible = .FALSE.}.

\itt{getdua} is a scalar variable of type default \logical, that
must be set \true\ if the user-provided estimates of the dual variables
should be replaced by estimates whose aim is to try to balance the
requirements of dual feasibility and complementary slackness,
and \false\ if users estimates are to be used.
The default is {\tt getdua = .FALSE.}.

\itt{puiseux} is a scalar variable of type default \logical, that
must be set \true\
if a Puiseux series will be used when extrapolating along the central path
and \false\ if a Taylor series is preferred.
We recommend using the Puiseux series unless the solution is known
to be non-degenerate.
The default is {\tt puiseux = .TRUE.}.

\itt{every\_order} is a scalar variable of type default \logical, that
must be set \true\
if every order of approximation up to {\tt series\_order} will be tried
and the best taken,
and \false\ if only the exact order specified by  {\tt series\_order}
will be used.
The default is {\tt every\_order = .TRUE.}.

\itt{feasol} is a scalar variable of type default \logical, that
should be set \true\
if the final solution obtained will be perturbed
so that variables close to their bounds are moved onto these bounds,
and \false\ otherwise.
The default is {\tt feasol = .FALSE.}.
%The default is {\tt feasol = .TRUE.}.

\itt{crossover} is a scalar variable of type default \logical,
that must be set \true\ if the solution is to be defined in terms
of linearly-independent constraints,
and  \false\ otherwise.
The default is {\tt crossover = .TRUE.}.

\itt{space\_critical} is a scalar variable of type default \logical,
that must be set \true\ if space is critical when allocating arrays
and  \false\ otherwise. The package may run faster if
{\tt space\_critical} is \false\ but at the possible expense of a larger
storage requirement. The default is {\tt space\_critical = .FALSE.}.

\itt{deallocate\_error\_fatal} is a scalar variable of type default \logical,
that must be set \true\ if the user wishes to terminate execution if
a deallocation  fails, and \false\ if an attempt to continue
will be made. The default is {\tt deallocate\_error\_fatal = .FALSE.}.

\itt{prefix} is a scalar variable of type default \character\
and length 30, that may be used to provide a user-selected
character string to preface every line of printed output.
Specifically, each line of output will be prefaced by the string
{\tt prefix(2:LEN(TRIM(prefix))-1)},
thus ignoring the first and last non-null components of the
supplied string. If the user does not want to preface lines by such
a string, they may use the default {\tt prefix = ""}.

\itt{FDC\_control} is a scalar variable of type
{\tt FDC\_control\_type}
whose components are used to control any detection of linear dependencies
performed by the package
{\tt \libraryname\_FDC}.
See the specification sheet for the package
{\tt \libraryname\_FDC}
for details, and appropriate default values.

\itt{SBLS\_control} is a scalar variable of type
{\tt SBLS\_control\_type}
whose components are used to control factorizations
performed by the package
{\tt \libraryname\_SBLS}.
See the specification sheet for the package
{\tt \libraryname\_SBLS}
for details, and appropriate default values.

\itt{FIT\_control} is a scalar variable of type
{\tt FIT\_control\_type}
whose components are used to control fitting of data to polynomials
performed by the package
{\tt \libraryname\_FIT}.
See the specification sheet for the package
{\tt \libraryname\_FIT}
for details, and appropriate default values.

\itt{ROOTS\_control} is a scalar variable of type
{\tt ROOTS\_control\_type}
whose components are used to control the polynomial root finding
performed by the package
{\tt \libraryname\_ROOTS}.
See the specification sheet for the package
{\tt \libraryname\_ROOTS}
for details, and appropriate default values.

\itt{CRO\_control} is a scalar variable of type
{\tt CRO\_control\_type}
whose components are used to control crossover
performed by the package
{\tt \libraryname\_CRO}.
See the specification sheet for the package
{\tt \libraryname\_CRO}
for details, and appropriate default values.

\end{description}
