\begin{description}

\itt{status} is a scalar variable of type default \integer, that gives the
exit status of the algorithm. 
%See Sections~\ref{galerrors} and \ref{galinfo}
See Section~\ref{galerrors} 
for details.

\itt{alloc\_status} is a scalar variable of type default \integer, that gives
the status of the last attempted array allocation or deallocation.
This will be 0 if {\tt status = 0}.

\itt{bad\_alloc} is a scalar variable of type default \character\
and length 80, that  gives the name of the last internal array 
for which there were allocation or deallocation errors.
This will be the null string if {\tt status = 0}. 

\itt{cg\_iter} is a scalar variable of type default \integer, that gives the
total number of conjugate-gradient iterations required.

\itt{factorization\_integer} is a scalar variable of type long
\integer, that gives the amount of integer storage used for the matrix 
factorization.

\itt{factorization\_real} is a scalar variable of type long \integer, 
that gives the amount of real storage used for the matrix factorization.

\ittf{obj} is a scalar variable of type default \realdp, that holds the
value of the objective function at the best estimate of the solution found.

\ittf{time} is a scalar variable of type {\tt \packagename\_time\_type} 
whose components are used to hold elapsed CPU and system clock times for the 
various parts of the calculation (see Section~\ref{typetime}).

\itt{FDC\_inform} is a scalar variable of type {\tt FDC\_inform\_type}
whose components are used to provide information about any detection of linear 
dependencies performed by the package {\tt \libraryname\_FDC}. See the
specification sheet for the package {\tt \libraryname\_FDC} for details.

\itt{SBLS\_inform} is a scalar variable of type {\tt SBLS\_inform\_type} 
whose components are used to hold information relating to the 
formation and factorization of the preconditioner. See the
documentation for the \galahad\ package {\tt SBLS} for further details.

\itt{GLTR\_inform} is a scalar variable of type {\tt GLTR\_inform\_type} 
whose components are used to hold information relating to the 
computation of the solution via the conjugate-gradient method. See the
documentation for the \galahad\ package {\tt GLTR} for further details.

\end{description}
