\documentclass{galahad}

% set the package name

\newcommand{\package}{gltr}
\newcommand{\packagename}{GL\-TR}
\newcommand{\fullpackagename}{\libraryname\_\packagename}

\begin{document}

\galheader

%%%%%%%%%%%%%%%%%%%%%% SUMMARY %%%%%%%%%%%%%%%%%%%%%%%%

\galsummary
Given real $n$ by $n$ symmetric matrices $\bmH$ and $\bmM$
(with $\bmM$ positive definite), a real
$n$ vector $\bmc$ and scalars $\Delta>0$ and $f_0$, this package finds an
{\bf approximate minimizer of the quadratic objective function
$\half \bmx^T \bmH \bmx  +  \bmc^T \bmx + f_0$, where the vector $\bmx$ is
required to satisfy the constraint $\|\bmx\|_{\bmM} \leq  \Delta$}, and
where the $\bmM$-norm of $\bmx$ is $\|\bmx\|_{\bmM} = \sqrt{\bmx^T \bmM \bmx}$.
This problem commonly occurs as a trust-region subproblem in nonlinear
optimization calculations.
The method may be suitable for large $n$ as no factorization of $\bmH$ is
required. Reverse communication is used to obtain
matrix-vector products of the form $\bmH \bmz$ and $\bmM^{-1} \bmz$.

\noindent
The package may also be used to solve the related problem in which $\bmx$ is
instead required to satisfy the
{\bf equality constraint $\|\bmx\|_{\bmM} = \Delta$}.

%%%%%%%%%%%%%%%%%%%%%% attributes %%%%%%%%%%%%%%%%%%%%%%%%

\galattributes
\galversions{\tt  \fullpackagename\_single, \fullpackagename\_double}.
\galuses
{\tt \libraryname\_SY\-M\-BOLS},
{\tt \libraryname\_SPACE}, {\tt \libraryname\_\-RAND},
{\tt \libraryname\_\-NORMS},
{\tt \libraryname\_ROOTS}, {\tt \libraryname\_SPECFILE},
{\tt *TTRF}.
\galdate April 1997.
\galorigin N. I. M. Gould, Rutherford Appleton Laboratory.
\gallanguage Fortran~95 + TR 15581 or Fortran~2003.

%%%%%%%%%%%%%%%%%%%%%% HOW TO USE %%%%%%%%%%%%%%%%%%%%%%%%

\galhowto

Access to the package requires a {\tt USE} statement such as

\medskip\noindent{\em Single precision version}

\hskip0.5in {\tt USE \fullpackagename\_single}

\medskip\noindent{\em Double precision version}

\hskip0.5in {\tt USE  \fullpackagename\_double}

\medskip

\noindent
If it is required to use both modules at the same time, the derived types
{\tt \packagename\_control\_type}, {\tt \packagename\_info\_type},
{\tt \packagename\_data\_type},
(Section~\ref{galtypes})
and the subroutines
{\tt \packagename\_initialize},
{\tt \packagename\_solve}, {\tt \packagename\_terminate}
(Section~\ref{galarguments})
and
{\tt \packagename\_read\_specfile}
(Section~\ref{galfeatures})
must be renamed on one of the {\tt USE} statements.

%%%%%%%%%%%%%%%%%%%%%% derived types %%%%%%%%%%%%%%%%%%%%%%%%

\galtypes
Three derived data types are accessible from the package.

%%%%%%%%%%% control type %%%%%%%%%%%

\subsubsection{The derived data type for holding control
 parameters}\label{typecontrol}
The derived data type
{\tt \packagename\_control\_type}
is used to hold controlling data. Default values may be obtained by calling
{\tt \packagename\_initialize}
(see Section~\ref{subinit}). The components of
{\tt \packagename\_control\_type}
are:

\begin{description}
\itt{error} is a scalar variable of type default \integer, that holds the
stream number for error messages.
Printing of error messages in
{\tt \packagename\_solve} and {\tt \packagename\_terminate}
is suppressed if ${\tt error} \leq {\tt 0}$.
The default is {\tt error = 6}.

\ittf{out} is a scalar variable of type default \integer, that holds the
stream number for informational messages.
Printing of informational messages in
{\tt \packagename\_solve} is suppressed if ${\tt out} < {\tt 0}$.
The default is {\tt out = - 1}.

\itt{print\_level} is a scalar variable of type default \integer,
that is used
to control the amount of informational output which is required. No
informational output will occur if ${\tt print\_level} \leq {\tt 0}$. If
{\tt print\_level = 1} a single line of output will be produced for each
iteration of the process. If {\tt print\_level} $\geq$ {\tt 2} this output
will be increased to provide significant detail of each iteration.
The default is {\tt print\_level = 0}.

\itt{itmax} is a scalar variable of type default \integer, that holds the
maximum number of iterations which will be allowed in
{\tt \packagename\_solve}.
If {\tt itmax} is set to a negative number, it will be reset by
{\tt \packagename\_solve} to $n$.
The default is {\tt itmax = -1}.

\itt{lanczos\_itmax} is a scalar variable of type default \integer,
that holds the
maximum number of iterations that may be performed when the iterates
encounter the boundary of the constraint.
If {\tt lanczos\_itmax} is set to a negative number, it will be reset by
{\tt \packagename\_solve} to $n$.
The default is {\tt lanczos\_itmax = -1}.

\itt{unitm} is a scalar variable of type default \logical, that
must be set \true\ if the matrix $\bmM$ is the identity matrix, and
\false\
otherwise.
The default is {\tt unitm = \true}.

\itt{extra\_vectors} is a scalar variable of type default \integer,
that specifies the number of additional vectors of length $n$
that will be allocated to try to speed up the computation if the
constraint boundary  is encountered.
The default is {\tt extra\_vectors = 0}.

\itt{steihaug\_toint} is a scalar variable of type default \logical,
which must be set \true\ if the algorithm is required to stop at
the first point on the boundary of the constraint that is encountered,
and \false\  if the constraint boundary is to be investigated
further.  Setting {\tt steihaug\_toint} to \true\ can reduce the
amount of computation at the expense of obtaining a poorer estimate of
the solution.
The default is {\tt steihaug\_toint = \false}.

\itt{boundary} is a scalar variable of type default \logical, that
may be set \true\ if the user believes that the solution will
occur on the constraint boundary, and \false\
otherwise. A correct setting of {\tt boundary} may reduce the amount of
computation performed.
The default is {\tt boundary = \false}.

\itt{equality\_problem} is a scalar variable of type default \logical, that
may be set \true\ if the user requires that the solution
occur on the constraint boundary (i.e., that the inequality constraint be
replaced by $\|\bmx\|_{\bmM} = \Delta$), and \false\ otherwise.
The default is {\tt equality\_problem = \false}.

\itt{space\_critical} is a scalar variable of type default \logical, that
may be set \true\ if the user wishes the package to allocate as little
internal storage as possible, and \false\ otherwise. The package may
be more efficient if {\tt space\_critical} is set \false.
The default is {\tt space\_critical = \false}.

\itt{deallocate\_error\_fatal} is a scalar variable of type default \logical,
that may be set \true\ if the user wishes the package to return to the user
in the unlikely event that an internal array deallocation fails,
and \false\ if the package should be allowed to try to continue.
The default is {\tt deallocate\_error\_fatal = \false}.

\itt{stop\_relative} and {\tt stop\_absolute} are scalar variables of type
\realdp, that holds the
relative and absolute convergence tolerances (see Section~\ref{galmethod}).
The computed solution $\bmx$ is accepted by {\tt \packagename\_solve}
if the computed value of
$\|\bmH \bmx + \lambda \bmM \bmx + \bmc\|_{{\bmM^{-1}}} $ is less than or equal
to $\max (\| \bmc\|_{\bmM^{-1}} \ast$ {\tt stop\_relative},
{\tt stop\_absolute}$)$, where $\lambda$ is an estimate of the Lagrange
multiplier associated with the trust-region constraint.
The defaults are {\tt stop\_relative = $\sqrt{u}$} and
{\tt stop\_absolute = 0.0},
where $u$ is {\tt EPSILON(1.0)} ({\tt EPSILON(1.0D0)} in
{\tt \fullpackagename\_double}).

\itt{f\_min} is a scalar variable of type default
\realdp, that specifies the smallest value of the objective function
that is allowed without regarding the problem as being unbounded
from below. The default is {\tt f\_min = -HUGE(1.0)/2.0}
({\tt -HUGE(1.0D0)/2.0D0} in {\tt \fullpackagename\_double})

\itt{fraction\_opt} is a scalar variable of type default
\realdp, that specifies the fraction
of the optimal value which is to be considered acceptable by the algorithm.
A negative value is considered to be zero, and a value of larger than one
is considered to be one. Reducing {\tt fraction\_opt} below one will result
in a reduction of the computation performed at the expense of an inferior
optimal value.
The default is {\tt fraction\_opt = 1.0}.

\itt{f\_0} is a scalar variable of type default
\realdp, that gives the value of the constant term $f_0$ in the
quadratic objective function. This value has no effect on the
computed minimizer $\bmx$. The default is {\tt f\_0 = 0.0}.

\itt{rminvr\_zero} is a scalar variable of type default
\realdp, that gives the smallest value that the square of the $\bmM$-norm of the
gradient of the objective function may be before it is considered to be zero.
The default is {\tt rminvr\_zero = 10 $u$},
where $u$ is {\tt EPSILON(1.0)} ({\tt EPSILON(1.0D0)} in
{\tt \fullpackagename\_double}).

\itt{prefix} is a scalar variable of type default \character\
and length 30, that may be used to provide a user-selected
character string to preface every line of printed output.
Specifically, each line of output will be prefaced by the string
{\tt prefix(2:LEN(TRIM(prefix))-1)},
thus ignoring the first and last non-null components of the
supplied string. If the user does not want to preface lines by such
a string, they may use the default {\tt prefix = ""}.

\end{description}

%%%%%%%%%%% info type %%%%%%%%%%%

\subsubsection{The derived data type for holding informational
 parameters}\label{typeinfo}
The derived data type
{\tt \packagename\_info\_type}
is used to hold parameters that give information about the progress and needs
of the algorithm. The components of
{\tt \packagename\_info\_type}
are:

\begin{description}
\itt{status} is a scalar variable of type default \integer, that gives the
current status of the algorithm. See Sections~\ref{galreverse} and
\ref{galerrors} for details.

\itt{alloc\_status} is a scalar variable of type default \integer,
that gives the status of the last internal array allocation
or deallocation. This will be 0 if {\tt status = 0}.

\itt{bad\_alloc} is a scalar variable of type default \character\
and length 80, that  gives the name of the last internal array
for which there were allocation or deallocation errors.
This will be the null string if {\tt status = 0}.

\itt{multiplier} is a scalar variable of type default
\realdp, that holds the
value of the Lagrange multiplier associated with the constraint.

\itt{mnormx} is a scalar variable of type default
\realdp,
that holds the current
value of $\|\bmx\|_{\bmM} $.

\itt{leftmost} is a scalar variable of type default
\realdp, that holds an estimate of the leftmost eigenvalue of the
matrix pencil $(\bmH,\bmM)$.

\ittf{iter} is a scalar variable of type default \integer, that holds the
current number of Lanczos vectors used.

\itt{iter\_pass2} is a scalar variable of type default \integer, that holds the
current number of Lanczos vectors used in the second pass.

\itt{negative\_curvature} is a scalar variable of type default \logical, that
is set \true\ if $\bmH$ has been found to be indefinite during the calculation
and \false\ otherwise.
\end{description}

%%%%%%%%%%% data type %%%%%%%%%%%

\subsubsection{The derived data type for holding problem data}\label{typedata}
The derived data type
{\tt \packagename\_data\_type}
is used to hold all the data for a particular problem between calls of
{\tt \packagename} procedures.
This data should be preserved, untouched, from the initial call to
{\tt \packagename\_initialize}
to the final call to
{\tt \packagename\_terminate}.

%%%%%%%%%%%%%%%%%%%%%% argument lists %%%%%%%%%%%%%%%%%%%%%%%%

\galarguments
There are three procedures for user calls
(see Section~\ref{galfeatures} for further features):

\begin{enumerate}
\item The subroutine
      {\tt \packagename\_initialize}
      is used to set default values, and initialize private data.
\item The subroutine
      {\tt \packagename\_solve}
      is called repeatedly to solve the problem.
      On each exit, the user may be expected to provide additional
      information and, if necessary, re-enter the subroutine.
\item The subroutine
      {\tt \packagename\_terminate}
      is provided to allow the user to automatically deallocate array
       components of the private data, allocated by
       {\tt \packagename\_solve},
       at the end of the solution process.
       It is important to do this if the data object is re-used for another
       problem since {\tt \packagename\_initialize} cannot test for this
       situation,
       and any existing associated targets will subsequently become
       unreachable.
\end{enumerate}
We use square brackets {\tt [ ]} to indicate \optional\ arguments.

%%%%%% initialization subroutine %%%%%%

\subsubsection{The initialization subroutine}\label{subinit}
 Default values are provided as follows:

\hskip0.5in
{\tt CALL \packagename\_initialize( data, control, inform )}

\begin{description}

\itt{data} is a scalar \intentinout argument of type
{\tt \packagename\_data\_type}
(see Section~\ref{typedata}). It is used to hold data about the problem being
solved.

\itt{control} is a scalar \intentout argument of type
{\tt \packagename\_control\_type}
(see Section~\ref{typecontrol}).
On exit, {\tt control} contains default values for the components as
described in Section~\ref{typecontrol}.
These values should only be changed after calling
{\tt \packagename\_initialize}.

\itt{inform} is a scalar \intentout\ argument of type
{\tt \packagename\_info\_type}
(see Section~\ref{typeinfo}). A successful call to
{\tt \packagename\_initialize}
is indicated when the  component {\tt status} has the value 0.
For other return values of {\tt status}, see Section~\ref{galerrors}.

\end{description}

%%%%%%%%% problem solution subroutine %%%%%%

\subsubsection{The optimization problem solution subroutine}
The optimization problem solution algorithm is called as follows:

\hskip0.5in
{\tt CALL \packagename\_solve( n, radius, f, X, R, VECTOR, data, control, inform )}

\begin{description}

\ittf{n} is a scalar \intentin\ argument of type default \integer, that must be
set to the number of unknowns, $n$. {\bf Restriction: } $n  >  0$.

\itt{radius} is a scalar \intentin\ variable of type default
\realdp,
that must be set on initial entry
to the value of the radius of the quadratic constraint, $\Delta$.
{\bf Restriction: } $\Delta > 0$.

\ittf{f} is a scalar \intentinout\ variable of type default
\realdp,
that holds an estimate of
the optimal value of the quadratic objective function.
On initial entry, {\tt f} need not be set.
It must not be changed between entries.
On exit, {\tt f} contains the current best estimate of the optimal objective
value.

\ittf{X} is an array \intentinout\ argument of dimension {\tt n} and
type default \realdp,
that holds an estimate of the solution $\bmx$ of the linear system.
On initial entry, {\tt X} need not be set.
It must not be changed between entries.
On exit, {\tt X} contains the current best estimate of the solution.

\ittf{R} is an array \intentinout\ argument of dimension {\tt n} and
type default \realdp,
that is used to hold the gradient $\bmH \bmx  +  \bmc$ of the objective
function at the current estimate of the solution.
On initial entry, {\tt R} must contain the vector $\bmc$.
If {\tt inform\%status} = 5 on exit, it must be reset to  $\bmc$; otherwise
it must be left unchanged. On exit, {\tt R} contains the gradient of the objective
function at the current best estimate of the solution.

\itt{VECTOR} is an array \intentinout\ argument of dimension {\tt n}
and type default \realdp,
that is used to pass information from and to {\tt \packagename\_solve},
as explained in Section~\ref{galreverse}.
On initial entry, {\tt VECTOR} need not be set. On exit, the actual content of
the array depends on the value of the parameter {\tt inform\%status}
(see Section~\ref{galreverse}).

\itt{data} is a scalar \intentinout argument of type
{\tt \packagename\_data\_type}
(see Section~\ref{typedata}). It is used to hold data about the problem being
solved. It must not have been altered {\bf by the user} since the last call to
{\tt \packagename\_initialize}.

\itt{control} is a scalar \intentin\ argument of type
{\tt \packagename\_control\_type}.
(see Section~\ref{typecontrol}).
Default values may be assigned by calling {\tt \packagename\_initialize}
prior to the first call to {\tt \packagename\_solve}.

\itt{inform} is a scalar \intentinout argument of type
{\tt \packagename\_info\_type}
(see Section~\ref{typeinfo}).
On initial entry, the component {\tt status} must be set to {\tt 1.}
The remaining components need not be set.
A successful call to
{\tt \packagename\_solve}
is indicated when the  component {\tt status} has the value 0.
For other return values of {\tt status}, see Sections~\ref{galreverse}
and \ref{galerrors}.
\end{description}

%%%%%%% termination subroutine %%%%%%

\subsubsection{The  termination subroutine}
All previously allocated arrays are deallocated as follows:

\hskip0.5in
{\tt CALL \packagename\_terminate( data, control, inform )}

\begin{description}

\itt{data} is a scalar \intentinout argument of type
{\tt \packagename\_data\_type}
exactly as for
{\tt \packagename\_solve}
that must not have been altered {\bf by the user} since the last call to
{\tt \packagename\_initialize}.
On exit, array components will have been deallocated.

\itt{control} is a scalar \intentin argument of type
{\tt \packagename\_control\_type}
exactly as for
{\tt \packagename\_solve}.

\itt{inform} is a scalar \intentout argument of type
{\tt \packagename\_info\_type}
exactly as for
{\tt \packagename\_solve}.
Only the component {\tt status} will be set on exit, and a
successful call to
{\tt \packagename\_terminate}
is indicated when this  component {\tt status} has the value 0.
For other return values of {\tt status}, see Section~\ref{galerrors}.

\end{description}

%%%%%%%%%%%%%%%%%%%%%% Reverse communications %%%%%%%%%%%%%%%%%%%%%%%%

\galreverse
A positive value of {\tt inform\%status} on exit from
{\tt \packagename\_solve} indicates that the user needs to take appropriate
action before re-entering the subroutine. Possible values are:

\begin{description}

\itt{2.} The user must perform the preconditioning operation
\disp{\bmy := \bmM^{-1} \bmz,}
and recall {\tt \packagename\_solve}.
The vector $\bmz$ is available in the array {\tt VECTOR}, and the result
$\bmy$ must be placed in {\tt VECTOR}.
No argument except {\tt VECTOR} should be altered before recalling
{\tt \packagename\_solve}. This return can only occur when
{\tt control\%unitm} is \false.

\itt{3.} The user must perform the matrix-vector product
\disp{\bmy := \bmH \bmz}
and recall {\tt \packagename\_solve}.
The vector $\bmz$ is available in the array {\tt VECTOR}, and the result
$\bmy$ must be placed in {\tt VECTOR}.
No argument except {\tt VECTOR} should be altered before recalling
{\tt \packagename\_solve}.

\itt{5.} The user should reset {\tt R} to $\bmc$ and recall
{\tt \packagename\_solve}.
No argument except {\tt R} should be altered before recalling
{\tt \packagename\_solve}.
\end{description}

%%%%%%%%%%%%%%%%%%%%%% Warning and error messages %%%%%%%%%%%%%%%%%%%%%%%%

\galerrors
A negative value of  {\tt inform\%status} on exit from
{\tt \packagename\_solve}
or
{\tt \packagename\_terminate}
indicates that an error might have occurred. No further calls should be made
until the error has been corrected. Possible values are:

\begin{description}
\itt{\galerrallocate.} An allocation error occurred. A message indicating
the offending
array is written on unit {\tt control\%error}, and the returned allocation
status and a string containing the name of the offending array
are held in {\tt inform\%alloc\_\-status}
and {\tt inform\%bad\_alloc} respectively.

\itt{\galerrdeallocate.} A deallocation error occurred.
A message indicating the offending
array is written on unit {\tt control\%error} and the returned allocation
status and a string containing the name of the offending array
are held in {\tt inform\%alloc\_\-status}
and {\tt inform\%bad\_alloc} respectively.

\itt{\galerrrestrictions.} ({\tt \packagename\_solve} only)
One of the restrictions
{\tt n > 0}
or
{\tt radius > 0}
has been violated.

\itt{\galerrpreconditioner.} ({\tt \packagename\_solve} only)
The matrix $\bmM$ appears not to be positive definite.

\itt{\galerrmaxiterations.} ({\tt \packagename\_solve} only) More than
{\tt control\%itmax} iterations have been performed without obtaining
convergence.

\itt{\galwarnboundary.} ({\tt \packagename\_solve} only)
The constraint boundary has been
encountered when the input value of {\tt control\%steihaug\-\_\-toint} was
set \true. The
solution is unlikely to have achieved the accuracy required by
{\tt control\%stop\_rela\-tive} and {\tt control\%stop\_absolute}.

\itt{\galerrfmin} ({\tt \packagename\_solve} only)
An objective function value smaller than {\tt control\%f\_min} has
been detected.

\end{description}

%%%%%%%%%%%%%%%%%%%%%% Re-entry %%%%%%%%%%%%%%%%%%%%%%%%

\subsection{Re-entry with a new value of $\Delta$}
It commonly happens that, having solved the problem for a particular value
of the radius $\Delta$, a user now wishes to solve the problem for a different
value of $\Delta$. Rather than restarting the calculation with
{\tt inform\%status = 1,} a useful approximation may be found
resetting {\tt radius} to the new required value and {\tt R} to $\bmc$,
and recalling {\tt \packagename\_solve}
with {\tt inform\%status = 4} and the remaining arguments unchanged.
This will determine the best solution within the Krylov space investigated
in the previous minimization (see Section~\ref{galmethod}).

%%%%%%%%%%%%%%%%%%%%%% Further features %%%%%%%%%%%%%%%%%%%%%%%%

\galfeatures
\noindent In this section, we describe an alternative means of setting
control parameters, that is components of the variable {\tt control} of type
{\tt \packagename\_control\_type}
(see Section~\ref{typecontrol}),
by reading an appropriate data specification file using the
subroutine {\tt \packagename\_read\_specfile}. This facility
is useful as it allows a user to change  {\tt \packagename} control parameters
without editing and recompiling programs that call {\tt \packagename}.

A specification file, or specfile, is a data file containing a number of
"specification commands". Each command occurs on a separate line,
and comprises a "keyword",
which is a string (in a close-to-natural language) used to identify a
control parameter, and
an (optional) "value", which defines the value to be assigned to the given
control parameter. All keywords and values are case insensitive,
keywords may be preceded by one or more blanks but
values must not contain blanks, and
each value must be separated from its keyword by at least one blank.
Values must not contain more than 30 characters, and
each line of the specfile is limited to 80 characters,
including the blanks separating keyword and value.

The portion of the specification file used by
{\tt \packagename\_read\_specfile}
must start
with a "{\tt BEGIN \packagename}" command and end with an
"{\tt END}" command.  The syntax of the specfile is thus defined as follows:
\begin{verbatim}
  ( .. lines ignored by GLTR_read_specfile .. )
    BEGIN GLTR
       keyword    value
       .......    .....
       keyword    value
    END
  ( .. lines ignored by GLTR_read_specfile .. )
\end{verbatim}
where keyword and value are two strings separated by (at least) one blank.
The ``{\tt BEGIN \packagename}'' and ``{\tt END}'' delimiter command lines
may contain additional (trailing) strings so long as such strings are
separated by one or more blanks, so that lines such as
\begin{verbatim}
    BEGIN GLTR SPECIFICATION
\end{verbatim}
and
\begin{verbatim}
    END GLTR SPECIFICATION
\end{verbatim}
are acceptable. Furthermore,
between the
``{\tt BEGIN \packagename}'' and ``{\tt END}'' delimiters,
specification commands may occur in any order.  Blank lines and
lines whose first non-blank character is {\tt !} or {\tt *} are ignored.
The content
of a line after a {\tt !} or {\tt *} character is also
ignored (as is the {\tt !} or {\tt *}
character itself). This provides an easy manner to "comment out" some
specification commands, or to comment specific values
of certain control parameters.

The value of a control parameters may be of three different types, namely
integer, logical or real.
Integer and real values may be expressed in any relevant Fortran integer and
floating-point formats (respectively). Permitted values for logical
parameters are "{\tt ON}", "{\tt TRUE}", "{\tt .TRUE.}", "{\tt T}",
"{\tt YES}", "{\tt Y}", or "{\tt OFF}", "{\tt NO}",
"{\tt N}", "{\tt FALSE}", "{\tt .FALSE.}" and "{\tt F}".
Empty values are also allowed for
logical control parameters, and are interpreted as "{\tt TRUE}".

The specification file must be open for
input when {\tt \packagename\_read\_specfile}
is called, and the associated device number
passed to the routine in device (see below).
Note that the corresponding
file is {\tt REWIND}ed, which makes it possible to combine the specifications
for more than one program/routine.  For the same reason, the file is not
closed by {\tt \packagename\_read\_specfile}.

\subsubsection{To read control parameters from a specification file}
\label{readspec}

Control parameters may be read from a file as follows:
\hskip0.5in
\def\baselinestretch{0.8} {\tt \begin{verbatim}
     CALL GLTR_read_specfile( control, device )
\end{verbatim}}
\def\baselinestretch{1.0}

\begin{description}
\itt{control} is a scalar \intentinout argument of type
{\tt \packagename\_control\_type}
(see Section~\ref{typecontrol}).
Default values should have already been set, perhaps by calling
{\tt \packagename\_initialize}.
On exit, individual components of {\tt control} may have been changed
according to the commands found in the specfile. Specfile commands and
the component (see Section~\ref{typecontrol}) of {\tt control}
that each affects are given in Table~\ref{specfile}.

\bctable{|l|l|l|}
\hline
  command & component of {\tt control} & value type \\
\hline
  {\tt error-printout-device} & {\tt \%error} & integer \\
  {\tt printout-device} & {\tt \%out} & integer \\
  {\tt print-level} & {\tt \%print\_level} & integer \\
  {\tt maximum-number-of-iterations} & {\tt \%itmax} & integer \\
  {\tt maximum-number-of-Lanczos-iterations} & {\tt \%lanczos\_itmax} & integer \\
  {\tt number-extra-n-vectors-used} & {\tt \%extra\_vectors} & integer \\
  {\tt relative-accuracy-required} & {\tt \%stop\_relative} & real \\
  {\tt absolute-accuracy-required} & {\tt \%stop\_absolute} & real \\
  {\tt small-f-stop} & {\tt \%f\_min} & real \\
  {\tt fraction-optimality-required} & {\tt \%fraction\_opt} & real \\
  {\tt constant-term-in-objective} & {\tt \%f\_0} & real \\
  {\tt zero-gradient-tolerance} & {\tt \%rminvr\_zero} & real \\

  {\tt two-norm-trust-region} & {\tt \%unitm} & logical \\
  {\tt stop-as-soon-as-boundary-encountered} & {\tt \%steihaug\_toint} & logical \\
  {\tt solution-is-likely-on-boundary} & {\tt \%boundary} & logical \\
  {\tt equality-problem} & {\tt \%equality\_problem} & logical \\
  {\tt space-critical} & {\tt \%space\_critical} & logical \\
  {\tt deallocate-error-fatal} & {\tt \%deallocate\_error\_fatal} & logical \\
\hline

\ectable{\label{specfile}Specfile commands and associated
components of {\tt control}.}

\itt{device} is a scalar \intentin argument of type default \integer,
that must be set to the unit number on which the specfile
has been opened. If {\tt device} is not open, {\tt control} will
not be altered and execution will continue, but an error message
will be printed on unit {\tt control\%error}.

\end{description}

%%%%%%%%%%%%%%%%%%%%%% Information printed %%%%%%%%%%%%%%%%%%%%%%%%

\galinfo
If {\tt control\%print\_level} is positive, information about the progress
of the algorithm will be printed on unit {\tt control\-\%\-out}.
If {\tt control\%print\_level = 1}, a single line of output will be produced
for each iteration of the process. So long as the current estimate lies
within the constraint boundary, this will include
the iteration number, the value of the objective function, the
norm of the gradient, the step taken, the two-norm of the latest search
direction, the norm $\|\bmx\|_{\bmM}$, and the curvature of $\bmH$ observed
along the current search direction (see Section~\ref{galmethod}).
A further message will be printed
if the constraint boundary is encountered during the current iteration.
Thereafter, the one-line summary will record
the iteration number, the value of the objective function, the
norm of the gradient of the Lagrangian,
$\|\bmH \bmx  +  \lambda \bmM \bmx  +  \bmc\|_{\bmM^{-1}}$,
the value of the Lagrange multiplier, the
number of iterations taken to solve the ``tridiagonal'' subproblem
and the exit status from this calculation (once again, see
Section~\ref{galmethod}).
If {\tt control\%print\_level} $\geq$ {\tt 2}, this
output will be increased to provide significant detail of each iteration.
This extra output includes a complete history of the inner iteration required
to solve the ``tridiagonal'' subproblem, and records the leftmost eigenvalue
of $\bmT_k$ (if needed), along with the estimates of the Lagrange
multipliers and the norm $\|\bmx\|_{\bmM}$
generated during the inner iteration.

%%%%%%%%%%%%%%%%%%%%%% GENERAL INFORMATION %%%%%%%%%%%%%%%%%%%%%%%%

\galgeneral

\galcommon None.
\galworkspace Provided automatically by the module.
\galroutines {\tt \packagename\_solve} calls the LAPACK subroutine
{\tt *PTTRF}, where {\tt *} is {\tt S} for
the default real version and {\tt D} for the double precision version.
\galmodules {\tt \packagename\_solve} calls the \galahad\ packages
{\tt GALAHAD\_SY\-M\-BOLS},
{\tt \libraryname\_\-SPACE},
{\tt \libraryname\-\_\-RAND},
{\tt \libraryname\_\-NORMS},
{\tt \libraryname\_\-ROOTS} and
{\tt GALAHAD\_SPECFILE}.
\galio Output is under control of the arguments
{\tt control\%error}, {\tt control\%out} and {\tt control\%print\_level}.
\galrestrictions $n  >  0, \;  \Delta  >  0$.
\galportability ISO Fortran~95 + TR 15581 or Fortran~2003.
The package is thread-safe.

%%%%%%%%%%%%%%%%%%%%%% METHOD %%%%%%%%%%%%%%%%%%%%%%%%

\galmethod
The required solution $\bmx$ necessarily satisfies the optimality condition
$\bmH \bmx + \lambda \bmM \bmx + \bmc = 0$, where $\lambda \geq 0$
is a Lagrange
multiplier corresponding to the constraint $\|\bmx\|_{\bmM}  \leq  \Delta$.
In addition, the matrix $\bmH + \lambda \bmM$ will be positive definite.

\noindent
The method is iterative. Starting  with the vector $\bmM^{-1} \bmc$,
a matrix of Lanczos vectors is built one column at a time
so that the $k$-th column is generated during
iteration $k$. These columns span a so-called Krylov space.
The resulting $n$ by $k$ matrix $\bmQ_k $ has the
property that $\bmQ_{k}^T \bmH \bmQ_k^{}  =  \bmT_{k}^{}$,
where $\bmT_k$ is tridiagonal. An approximation to the
required solution may then be expressed formally as
\disp{\bmx_{k+1}  =  \bmQ_k \bmy_k}
where $\bmy_k $ solves the ``tridiagonal'' subproblem of minimizing
\eqn{4.1}{\half  \bmy^T \bmT_k \bmy  + \| \bmc\|_{\bmM^{-1} } \bme_{1}^T \bmy
 \tim{subject to the constraint} \|\bmy\|_2  \leq  \Delta ,}
and where $\bme_1$ is the first unit vector.

If the solution to \req{4.1} lies interior to the constraint, the required
solution $\bmx_{k+1}$ may simply be found as the $k$-th (preconditioned)
conjugate-gradient iterate. This solution can be obtained without the need to
access the whole matrix $\bmQ_k$.
These conjugate-gradient iterates increase in $\bmM$-norm, and
thus once one of them exceeds $\Delta$ in $\bmM$-norm, the solution must occur
on the constraint boundary. Thereafter, the solution to \req{4.1} is less
easy to obtain, but an efficient inner iteration to solve \req{4.1} is
none the less possible because $\bmT_k $ is tridiagonal.
It is possible to measure the optimality measure
$\|\bmH \bmx  +  \lambda \bmM \bmx  +  \bmc\|_{\bmM^{-1}}$
without computing $\bmx_{k+1}$, and thus without
needing $\bmQ_k $. Once this measure is sufficiently small, a second pass
is required to obtain the estimate $\bmx_{k+1} $ from $\bmy_k $.
As this second pass is an additional expense, a record is kept of the
optimal objective function values for each value of $k$, and the second
pass is only performed so far as to ensure a given fraction of the
final optimal objective value. Large savings may be made in the second
pass by choosing the required fraction to be significantly smaller than one.

A cheaper alternative is to use the Steihuag-Toint strategy, which is simply
to stop at the first boundary point encountered along the piecewise
linear path generated by the conjugate-gradient iterates. Note that if
$\bmH$ is significantly indefinite, this strategy often produces a far from
optimal point, but is effective when $\bmH$ is positive definite or almost
so.

\galreference
The method is described in detail in
\vspace*{1mm}

\noindent
N. I. M. Gould, S. Lucidi, M. Roma and Ph. L. Toint,
Solving the trust-region subproblem using the Lanczos method.
SIAM Journal on Optimization {\bf 9:2 } (1999), 504-525.

%%%%%%%%%%%%%%%%%%%%%% EXAMPLE %%%%%%%%%%%%%%%%%%%%%%%%

\galexample
Suppose we wish to solve a problem in 10,000 unknowns, whose data is
\disp{\bmH  = \mat{ccccc}{ -2 & 1 \\ 1 & -2 & . \\ & . & . & . \\ & & . & -2 & 1
                           \\ & & & 1 & -2 }, \;\;
\bmM  = \mat{ccccc}{ 2 &   \\   & 2  &   \\ & & . \\ & & & 2 & \\ & & & & 2 }
\tim{and}
\bmc  =  \vect{ 1 \\ 1 \\ . \\ 1 \\ 1 },}
with a radius $\Delta = 10$. Then we may use the following code

{\tt \small
\VerbatimInput{\packageexample}
}
\noindent
This produces the following output:
{\tt \small
\VerbatimInput{\packageresults}
}


\end{document}
