\begin{description}

\itt{status} is a scalar variable of type default \integer, that gives the
exit status of the algorithm.
%See Sections~\ref{galerrors} and \ref{galinfo}
See Section~\ref{galerrors}
for details.

\itt{alloc\_status} is a scalar variable of type default \integer, that gives
the status of the last attempted array allocation or deallocation.
This will be 0 if {\tt status = 0}.

\itt{bad\_alloc} is a scalar variable of type default \character\
and length 80, that  gives the name of the last internal array
for which there were allocation or deallocation errors.
This will be the null string if {\tt status = 0}.

\itt{iter} is a scalar variable of type default \integer, that
gives the number of iterations performed.

\itt{la04\_job} is a scalar variable of type default \integer, that
gives the value of the argument {\tt job} returned from {\tt LA04}.
See the specification sheet,
{\tt http://www.hsl.rl.ac.uk/specs/la04.pdf},
for the package {\tt LA04} for details.

\itt{la04\_job\_info} is a scalar variable of type default \integer, that
gives the value of the argument {\tt RINFO(35)} returned from an
unsuccessful call to {\tt LA04}.
See the specification sheet,
{\tt http://www.hsl.rl.ac.uk/specs/la04.pdf},
for the package {\tt LA04} for details.

\ittf{obj} is a scalar variable of type default \realdp, that holds the
value of the objective function at the best estimate of the solution found.

\itt{primal\_infeasibility} is a scalar variable of type default \realdp,
that holds the norm of the violation of primal optimality
(see Section~\ref{typetime}) at the best estimate of the solution found.

\itt{feasible} is a scalar variable of type default \logical, that has the
value \true\ if the output value of $\bmx$ satisfies the constraints,
and the value \false\ otherwise.

\ittf{time} is a scalar variable of type {\tt \packagename\_time\_type}
whose components are used to hold elapsed CPU and system clock times for the
various parts of the calculation (see Section~\ref{typetime}).

\end{description}
