\begin{description}

\itt{status} is a scalar variable of type default \integer, that gives the
exit status of the algorithm. 
%See Sections~\ref{galerrors} and \ref{galinfo}
See Section~\ref{galerrors}
for details.

\itt{alloc\_status} is a scalar variable of type default \integer, that gives
the status of the last attempted array allocation or deallocation.
This will be 0 if {\tt status = 0}.

\itt{bad\_alloc} is a scalar variable of type default \character\
and length 80, that  gives the name of the last internal array 
for which there were allocation or deallocation errors.
This will be the null string if {\tt status = 0}. 

\itt{major\_iter} is a scalar variable of type default \integer, that gives the
total number of major iterations required.

\ittf{iter} is a scalar variable of type default \integer, that gives the
total number of iterations required.

\itt{cg\_iter} is a scalar variable of type default \integer, that gives the
total number of conjugate-gradient inner iterations required.

\itt{factorization\_status} is a scalar variable of type default \integer, that 
gives the return status from the matrix factorization.

\itt{factorization\_integer} is a scalar variable of type long
\integer, that gives the amount of integer storage used for the matrix 
factorization.

\itt{factorization\_real} is a scalar variable of type long \integer, 
that gives the amount of real storage used for the matrix factorization.

\itt{nfacts} is a scalar variable of type default \integer, that gives the
total number of factorizations performed.

\itt{nmods} is a scalar variable of type default \integer, that gives the
total number of factorizations which were modified to 
ensure that the matrix is an appropriate preconditioner. 

\itt{num\_g\_infeas} is a scalar variable of type default \integer, that 
gives the total number of general constraints that are violated.

\itt{num\_b\_infeas} is a scalar variable of type default \integer, that 
gives the total number of simple bound constraints that are violated.

\ittf{obj} is a scalar variable of type default \realdp, that holds the
value of the objective function at the best estimate of the solution found.

\itt{infeas\_g} is a scalar variable of type default \realdp, that holds the
value of the infeasibility $v_g(\bmx)$.

\itt{infeas\_b} is a scalar variable of type default \realdp, that holds the
value of the infeasibility $v_b(\bmx)$.

\itt{merit} is a scalar variable of type default \realdp, that holds the
value of the merit function $q(\bmx) + \rho_g v_g(\bmx) + \rho_b v_b(\bmx)$ 
at the best estimate of the solution found.

\ittf{time} is a scalar variable of type {\tt \packagename\_time\_type} 
whose components are used to hold elapsed CPU  and system clock 
times for the various parts of the calculation (see Section~\ref{typetime}).

\itt{SLS\_inform} is a scalar variable argument of type 
{\tt SLS\_inform\_type} that is used to pass information
concerning the progress of the external packages used to 
factorize relevant symmetric matrices that arise. 
See the documentation for the \galahad\ package {\tt SLS} for further details. 

\end{description}
