\documentclass{galahad}

% set the release and package names

\newcommand{\package}{sbls}
\newcommand{\packagename}{SBLS}
\newcommand{\fullpackagename}{\libraryname\_\packagename}

\begin{document}

\galheader

%%%%%%%%%%%%%%%%%%%%%% SUMMARY %%%%%%%%%%%%%%%%%%%%%%%%

\galsummary
Given a {\bf block, symmetric matrix}
\disp{ \bmK_{H} = \mat{cc}{ \bmH & \bmA^T \\ \bmA  & - \bmC },}
this package constructs a variety of {\bf preconditioners} of the form
\eqn{prec1}{ \bmK_{G} = \mat{cc}{ \bmG & \bmA^T \\ \bmA  & - \bmC }.}
Here, the leading-block matrix $\bmG$ is a suitably-chosen
approximation to $\bmH$; it may either be prescribed {\bf explicitly}, in
which case a symmetric indefinite factorization of $\bmK_{G}$
will be formed using the \galahad\ package {\tt SLS},
or {\bf implicitly}, by requiring certain sub-blocks of $\bmG$
be zero. In the latter case, a factorization of $\bmK_{G}$ will be
obtained implicitly (and more efficiently) without recourse to {\tt SLS}.
In particular, for implicit preconditioners, a reordering
\eqn{prec2}{ \bmK_{G} = \bmP
\mat{ccc}{ \bmG_{11}^{} & \bmG_{21}^T & \bmA_1^T \\
\bmG_{21}^{} & \bmG_{22}^{} & \bmA_2^T \\
\bmA_{1}^{} & \bmA_{2}^{} & - \bmC} \bmP^T}
involving a suitable permutation $\bmP$ will be found, for some
invertible sub-block (``basis'') $\bmA_1$ of the columns of $\bmA$;
the selection and factorization of $\bmA_1$ uses
the \galahad\ package {\tt ULS}.
Once the preconditioner has been constructed,
solutions to the preconditioning system
\eqn{ls}{ \mat{cc}{ \bmG & \bmA^T \\ \bmA  & - \bmC } \vect{ \bmx \\ \bmy }
 = \vect{\bma \\ \bmb}}
may be obtained by the package.
Full advantage is taken of any zero coefficients in the matrices $\bmH$,
$\bmA$ and $\bmC$.

%%%%%%%%%%%%%%%%%%%%%% attributes %%%%%%%%%%%%%%%%%%%%%%%%

\galattributes
\galversions{\tt  \fullpackagename\_single, \fullpackagename\_double}.
\galuses {\tt GALAHAD\_CLOCK},
{\tt GALAHAD\_\-SY\-M\-BOLS},
{\tt GALAHAD\-\_SPACE},
{\tt GALAHAD\_SMT},
{\tt GALAHAD\_QPT},
{\tt GALAHAD\_SLS},
{\tt GALAHAD\_ULS},
{\tt GALAHAD\_SPECFILE},
\galdate April 2006.
\galorigin H. S. Dollar and N. I. M. Gould,
Rutherford Appleton Laboratory.
\gallanguage Fortran~95 + TR 15581 or Fortran~2003.
\galparallelism Some options may use OpenMP and its runtime library.

%%%%%%%%%%%%%%%%%%%%%% HOW TO USE %%%%%%%%%%%%%%%%%%%%%%%%

\galhowto

%\subsection{Calling sequences}

Access to the package requires a {\tt USE} statement such as

\medskip\noindent{\em Single precision version}

\hspace{8mm} {\tt USE \fullpackagename\_single}

\medskip\noindent{\em Double precision version}

\hspace{8mm} {\tt USE  \fullpackagename\_double}

\medskip

\noindent
If it is required to use both modules at the same time, the derived types
{\tt SMT\_type},
{\tt QPT\_problem\_type},
{\tt \packagename\_time\_type},
{\tt \packagename\_control\_type},
{\tt \packagename\_inform\_type}
and
{\tt \packagename\_data\_type}
(\S\ref{galtypes})
and the subroutines
{\tt \packagename\_initialize},
{\tt \packagename\_\-solve},
{\tt \packagename\_terminate},
(\S\ref{galarguments})
and
{\tt \packagename\_read\_specfile}
(\S\ref{galfeatures})
must be renamed on one of the {\tt USE} statements.

%%%%%%%%%%%%%%%%%%%%%% matrix formats %%%%%%%%%%%%%%%%%%%%%%%%

\galmatrix
Each of the input matrices $\bmH$, $\bmA$ and $\bmC$
may be stored in a variety of input formats.

\subsubsection{Dense storage format}\label{dense}
The matrix $\bmA$ is stored as a compact
dense matrix by rows, that is, the values of the entries of each row in turn are
stored in order within an appropriate real one-dimensional array.
Component $n \ast (i-1) + j$ of the storage array {\tt A\%val} will hold the
value $a_{ij}$ for $i = 1, \ldots , m$, $j = 1, \ldots , n$.
Since $\bmH$ and $\bmC$ are symmetric, only the lower triangular parts
(that is the part $h_{ij}$ for $1 \leq j \leq i \leq n$ and
$c_{ij}$ for $1 \leq j \leq i \leq m$) need be held. In these cases
the lower triangle will be stored by rows, that is
component $i \ast (i-1)/2 + j$ of the storage array {\tt H\%val}
will hold the value $h_{ij}$ (and, by symmetry, $h_{ji}$)
for $1 \leq j \leq i \leq n$. Similarly
component $i \ast (i-1)/2 + j$ of the storage array {\tt C\%val}
will hold the value $c_{ij}$ (and, by symmetry, $c_{ji}$)
for $1 \leq j \leq i \leq m$.

\subsubsection{Sparse co-ordinate storage format}\label{coordinate}
Only the nonzero entries of the matrices are stored. For the
$l$-th entry of $\bmA$, its row index $i$, column index $j$
and value $a_{ij}$
are stored in the $l$-th components of the integer arrays {\tt A\%row},
{\tt A\%col} and real array {\tt A\%val}, respectively.
The order is unimportant, but the total
number of entries {\tt A\%ne} is also required.
The same scheme is applicable to $\bmH$ and $\bmC$
(thus, for $\bmH$, requiring integer arrays {\tt H\%row}, {\tt H\%col}, a real
array  {\tt H\%val} and an integer value {\tt H\%ne}),
except that only the entries in the lower triangle need be stored.

\subsubsection{Sparse row-wise storage format}\label{rowwise}
Again only the nonzero entries are stored, but this time
they are ordered so that those in row $i$ appear directly before those
in row $i+1$. For the $i$-th row of $\bmA$, the $i$-th component of a
integer array {\tt A\%ptr} holds the position of the first entry in this row,
while {\tt A\%ptr} $(m+1)$ holds the total number of entries plus one.
The column indices $j$ and values $a_{ij}$ of the entries in the $i$-th row
are stored in components
$l =$ {\tt A\%ptr}$(i)$, \ldots ,{\tt A\%ptr} $(i+1)-1$ of the
integer array {\tt A\%col}, and real array {\tt A\%val}, respectively.
The same scheme is applicable to
$\bmH$ and $\bmC$ (thus, for $\bmH$,
requiring integer arrays {\tt H\%ptr}, {\tt H\%col}, and
a real array {\tt H\%val}),
except that only the entries in the lower triangle need be stored.

For sparse matrices, this scheme almost always requires less storage than
its predecessor.

\subsubsection{Diagonal storage format}\label{diagonal}
If $\bmH$ is diagonal (i.e., $h_{ij} = 0$ for all $1 \leq i \neq j \leq n$)
only the diagonals entries $h_{ii}$, $1 \leq i \leq n$,  need be stored,
and the first $n$ components of the array {\tt H\%val} may be used for
the purpose. The same applies to $\bmC$, but
there is no sensible equivalent for the non-square $\bmA$.

\subsubsection{Scaled-identity storage format}\label{scaled}
If $\bmH$ is a scalar multiple $h$ of the identity matrix, $h \bmI$,
only the value $h$ needs be stored,
and the first component of the array {\tt H\%val}
 may be used for
the purpose. The same applies to $\bmC$, but
as before there is no sensible equivalent for the non-square $\bmA$.

\subsubsection{Identity storage format}\label{identity}
If $\bmH$ is the identity matrix, $\bmI$,
no numerical data need be stored. The same applies to $\bmC$, but
once again there is no sensible equivalent for the non-square $\bmA$.

\subsubsection{Zero storage format}\label{zero}
Finally, if $\bmH$ is the zero matrix, $\bmzero$,
no numerical data need be stored. The same applies to both $\bmA$ and $\bmC$.

\subsection{Integer kinds}\label{Integer kinds}
We use the term
long \integer\ to denote {\tt INTEGER\-(kind=long)}, where
{\tt long = selected\_int\_kind(18))}.

%%%%%%%%%%%%%%%%%%%%%% OpenMP usage %%%%%%%%%%%%%%%%%%%%%%%%

\subsection{OpenMP}
OpenMP may be used by the {\tt \fullpackagename} package to provide
parallelism for some solver options in shared memory environments.
See the documentatiuon for the \galahad\ package {\tt SLS} for more details.
To run in parallel, OpenMP
must be enabled at compilation time by using the correct compiler flag
(usually some variant of {\tt -openmp}).
The number of threads may be controlled at runtime
by setting the environment variable {\tt OMP\_NUM\_THREADS}.

\noindent
The code may be compiled and run in serial mode.

%%%%%%%%%%%%%%%%%%%%%% derived types %%%%%%%%%%%%%%%%%%%%%%%%

\galtypes
Six derived data types are accessible from the package.

%%%%%%%%%%% matrix data type %%%%%%%%%%%

\subsubsection{The derived data type for holding matrices}\label{typesmt}
The derived data type {\tt SMT\_TYPE} is used to hold the matrices $\bmH$,
$\bmA$ and $\bmC$. The components of {\tt SMT\_TYPE} used here are:

\begin{description}

\ittf{m} is a scalar component of type default \integer,
that holds the number of rows in the matrix.

\ittf{n} is a scalar component of type default \integer,
that holds the number of columns in the matrix.

\ittf{type} is a rank-one allocatable array of type default \character, that
is used to indicate the storage scheme used. If the dense storage scheme
(see \S\ref{dense}), is used,
the first five components of {\tt type} must contain the
string {\tt DENSE}.
For the sparse co-ordinate scheme (see \S\ref{coordinate}),
the first ten components of {\tt type} must contain the
string {\tt COORDINATE},
for the sparse row-wise storage scheme (see \S\ref{rowwise}),
the first fourteen components of {\tt type} must contain the
string {\tt SPARSE\_BY\_ROWS},
for the diagonal storage scheme (see \S\ref{diagonal}),
the first eight components of {\tt type} must contain the
string {\tt DIAGONAL}.
for the scaled-identity storage scheme (see \S\ref{scaled}),
the first fifteen components of {\tt type} must contain the
string {\tt SCALED\_IDENTITY}.
and for the identity storage scheme (see \S\ref{identity}),
the first eight components of {\tt type} must contain the
string {\tt IDENTITY}.
It is also permissible to set
the first four components of {\tt type} to the either of the strings
{\tt ZERO} or {\tt NONE}
in the case of matrix $\bmC$ to indicate that $\bmC = 0$.

For convenience, the procedure {\tt SMT\_put}
may be used to allocate sufficient space and insert the required keyword
into {\tt type}.
For example, if {\tt H} is of derived type {\tt SMT\_type}
and we wish to use the co-ordinate storage scheme, we may simply
%\vspace*{-2mm}
{\tt
\begin{verbatim}
        CALL SMT_put( H%type, 'COORDINATE', istat )
\end{verbatim}
}
%\vspace*{-4mm}
\noindent
See the documentation for the \galahad\ package {\tt SMT}
for further details on the use of {\tt SMT\_put}.

\ittf{ne} is a scalar variable of type default \integer, that
holds the number of matrix entries.

\ittf{val} is a rank-one allocatable array of type default \realdp\,
and dimension at least {\tt ne}, that holds the values of the entries.
Each pair of off-diagonal entries $h_{ij} = h_{ji}$ of a {\em symmetric}
matrix $\bmH$ is represented as a single entry
(see \S\ref{dense}--\ref{rowwise}); the same applies to $\bmC$.
Any duplicated entries that appear in the sparse
co-ordinate or row-wise schemes will be summed.
If the matrix is stored using the diagonal scheme (see \S\ref{diagonal}),
{\tt val} should be of length {\tt n}, and the value of the {\tt i}-th
diagonal stored in {\tt val(i)}.
If the matrix is stored using the scaled-identity scheme (see \S\ref{identity}),
{\tt val(1)} should be set to $h$.

\ittf{row} is a rank-one allocatable array of type default \integer,
and dimension at least {\tt ne}, that may hold the row indices of the entries.
(see \S\ref{coordinate}).

\ittf{col} is a rank-one allocatable array of type default \integer,
and dimension at least {\tt ne}, that may hold the column indices of the entries
(see \S\ref{coordinate}--\ref{rowwise}).

\ittf{ptr} is a rank-one allocatable array of type default \integer,
and dimension at least {\tt m + 1}, that may hold the pointers to
the first entry in each row (see \S\ref{rowwise}).

\end{description}


%%%%%%%%%%% control type %%%%%%%%%%%

\subsubsection{The derived data type for holding control
 parameters}\label{typecontrol}
The derived data type
{\tt \packagename\_control\_type}
is used to hold controlling data. Default values may be obtained by calling
{\tt \packagename\_initialize}
(see \S\ref{subinit}),
while components may also be changed by calling
{\tt \fullpackagename\_read\-\_spec}
(see \S\ref{readspec}).
The components of
{\tt \packagename\_control\_type}
are:

\begin{description}

\itt{error} is a scalar variable of type default \integer, that holds the
stream number for error messages. Printing of error messages in
{\tt \packagename\_solve} and {\tt \packagename\_terminate}
is suppressed if {\tt error} $\leq 0$.
The default is {\tt error = 6}.

\ittf{out} is a scalar variable of type default \integer, that holds the
stream number for informational messages. Printing of informational messages in
{\tt \packagename\_solve} is suppressed if {\tt out} $< 0$.
The default is {\tt out = 6}.

\itt{print\_level} is a scalar variable of type default \integer, that is used
to control the amount of informational output which is required. No
informational output will occur if {\tt print\_level} $\leq 0$. If
{\tt print\_level} $= 1$, a single line of output will be produced for each
iteration of the process. If {\tt print\_level} $\geq 2$, this output will be
increased to provide significant detail of each iteration.
The default is {\tt print\_level = 0}.

\itt{new\_h} is a scalar variable of type default \integer, that is used
to indicate how $\bmH$ has changed (if at all) since the previous call
to {\tt SBLS\_form\_and\_factorize}. Possible values are:
\begin{description}
\itt{0} $\bmH$ is unchanged
\itt{1} the values in $\bmH$ have changed, but its nonzero structure
is as before.
\itt{2} both the values and structure of $\bmH$ have changed.
\end{description}
The default is {\tt new\_h = 2}.

\itt{new\_a} is a scalar variable of type default \integer, that is used
to indicate how $\bmA$ has changed (if at all) since the previous call
to {\tt SBLS\_form\_and\_factorize}. Possible values are:
\begin{description}
\itt{0} $\bmA$ is unchanged
\itt{1} the values in $\bmA$ have changed, but its nonzero structure
is as before.
\itt{2} both the values and structure of $\bmA$ have changed.
\end{description}
The default is {\tt new\_a = 2}.

\itt{new\_c} is a scalar variable of type default \integer, that is used
to indicate how $\bmC$ has changed (if at all) since the previous call
to {\tt SBLS\_form\_and\_factorize}. Possible values are:
\begin{description}
\itt{0} $\bmC$ is unchanged
\itt{1} the values in $\bmC$ have changed, but its nonzero structure
is as before.
\itt{2} both the values and structure of $\bmC$ have changed.
\end{description}
The default is {\tt new\_c = 2}.

\itt{preconditioner} is a scalar variable of type default \integer,
that specifies which preconditioner to be used;
positive values correspond to explicit-factorization preconditioners
while negative values indicate implicit-factorization ones. Possible values are:

\begin{description}
\itt{0} the preconditioner is chosen automatically on the basis of which option
        looks most likely to be the most efficient.
\itt{1} $\bmG$ is chosen to be the identity matrix.
\itt{2} $\bmG$ is chosen to be $\bmH$
\itt{3} $\bmG$ is chosen to be the diagonal matrix whose diagonals
        are the larger of those of $\bmH$ and a positive constant
        (see {\tt min\_diagonal} below).
\itt{4} $\bmG$ is chosen to be the band matrix of given semi-bandwidth
        whose entries coincide with those of $\bmH$ within the band.
        (see {\tt semi\_bandwidth} below).
\itt{5} $\bmG$ is chosen to be a diagonal matrix $\bmD$ specified by the
        user.
\itt{11} $\bmG$ is chosen so that $\bmG_{11} = 0$, $\bmG_{21} = 0$
        and $\bmG_{22} = \bmH_{22}$.
\itt{12} $\bmG$ is chosen so that $\bmG_{11} = 0$, $\bmG_{21} = \bmH_{21}$
        and $\bmG_{22} = \bmH_{22}$.
\itt{-1} for the special case when $\bmC = 0$,
        $\bmG$ is chosen so that $\bmG_{11} = 0$, $\bmG_{21} = 0$,
        $\bmG_{22}$ is the identity matrix, and the preconditioner is computed
        implicitly.
\itt{-2} for the special case when $\bmC = 0$,
        $\bmG$ is chosen so that $\bmG_{11} = 0$, $\bmG_{21} = 0$,
        $\bmG_{22} = \bmH_{22}$ and the preconditioner is computed
        implicitly.
\end{description}
Other values may be added in future.
The default is {\tt preconditioner = 0}.

\itt{semi\_bandwidth} is a scalar variable of type default \integer, that
specifies the semi-bandwidth of the band preconditioner when {\tt precon = 3},
if appropriate.
The default is {\tt semi\_bandwidth = 5}.

\itt{factorization} is a scalar variable of type default \integer, that
specifies which factorization of the preconditioner should be used.
Possible values are:

\begin{description}
\itt{0} the factorization is chosen automatically on the basis of which option
        looks likely to be the most efficient.
\itt{1} if $\bmG$ is diagonal and non-singular, a Schur-complement
        factorization, involving factors of $\bmG$ and $\bmA \bmG^{-1} \bmA^T$,
        will be used. Otherwise an augmented-system factorization, involving
        factors of $\bmK_G$, will be used.
\itt{2} an augmented-system factorization, involving factors of $\bmK_G$,
        will be used.
\itt{3} a null-space factorization (see \S\ref{galmethod}) will be used
provided that $\bmC = \bmzero$.
\end{description}
The default is {\tt factorization = 0}.

\itt{max\_col} is a scalar variable of type default \integer, that specifies
the maximum number of nonzeros in a column of $\bmA$ which is permitted
by the Schur-complement factorization.
The default is {\tt max\_col = 35}.

%\itt{indmin} is a scalar variable of type default \integer, that specifies
%an initial estimate as to the amount of integer workspace required by
%the factorization package {\tt SILS}.
%The default is {\tt indmin = 1000}.

%\itt{valmin} is a scalar variable of type default \integer, that specifies
%an initial estimate as to the amount of real workspace required by
%the factorization package {\tt SILS}.
%The default is {\tt valmin = 1000}.

%\itt{len\_ulsmin} is a scalar variable of type default \integer, that specifies
%an initial estimate as to the amount of workspace required by
%the factorization package {\tt ULS}.
%The default is {\tt len\_ulsmin = 1000}.

\itt{itref\_max} is a scalar variable of type default \integer, that specifies
the maximum number of iterative refinements allowed with each application
of the preconditioner.
The default is {\tt itref\_max = 1}.

%\itt{scaling} is a scalar variable of type default \integer, that is used

%\itt{ordering} is a scalar variable of type default \integer, that is used

%\itt{pivot\_tol}  is a scalar variable of type default
%\realdp, that holds the
%threshold pivot tolerance used by the matrix factorization.  See
%the documentation for the packages {\tt SILS} and {\tt ULS} for details.
%The default is {\tt pivot\_tol = 0.01}.

\itt{pivot\_tol\_for\_basis} is a scalar variable of type default
\realdp, that holds the relative
pivot tolerance used by the package {\tt ULS}
when computing the non-singular basis matrix $\bmA_1$ for
an implicit-factorization preconditioner. Since the calculation of a
suitable basis is crucial, it is sensible to pick a larger
value of {\tt pivot\_tol\_for\_basis} than
the default value {\tt relative\_pivot\_tolerance} provided by {\tt ULS}.
The default is {\tt pivot\_tol\_for\_basis = 0.5}.

%\itt{zero\_pivot} is a scalar variable of type default \realdp,
%that is used to detect singularity. Any pivot encountered during the
%factorization whose absolute value is less than or equal to  {\tt zero\_pivot}
%will be regarded as zero, and the matrix as singular.
%The default is {\tt zero\_pivot = EPSILON(1.0)}$^{0.75}$.

%\itt{static\_tolerance} is a scalar variable of type default \realdp,
%that is used

%\itt{static\_level} is a scalar variable of type default \realdp,
%that is used

\itt{min\_diagonal} is a scalar variable of type default \realdp,
that specifies the smallest permitted diagonal in $\bmG$ for some
of the preconditioners provided. See {\tt preconditioner} above.
The default is {\tt min\_diagonal = 0.00001}.

\itt{remove\_dependencies} is a scalar variable of type default \logical,
that must be set \true\ if linear dependent rows from the second
block equation $\bmA \bmx - \bmC \bmy = \bmb$ from \req{ls}
should be removed and \false\ otherwise.
The default is {\tt remove\_dependencies = .TRUE.}.

\itt{check\_basis} is a scalar variable of type default \logical,
that must be set \true\ if the basis matrix $\bmA_{1}^{}$ constructed
when using an implicit-factorization preconditioner or null-space factorization
should be assessed for ill conditioning and corrected if necessary. If these
precautions are not thought necessary, {\tt check\_basis} should be
set \false. The default is {\tt check\_basis = .TRUE.}.

\itt{find\_basis\_by\_transpose} is a scalar variable of type default \logical,
that must be set \true\ if the invertible sub-block $\bmA_1$ of the
columns of $\bmA$ is computed by analysing the transpose of $\bmA$
and \false\ if the analysis is based on $\bmA$ itself. Generally
an analysis based on the transpose is faster.
The default is {\tt find\_basis\_by\_transpose = .TRUE.}.

\itt{affine} is a scalar variable of type default \logical,
that must be set \true\ if the component $\bmb$ of the right-hand side
of the required system \req{ls} is zero, and \false\ otherwise.
Computational savings are possible when $\bmb = 0$.
The default is {\tt affine = .FALSE.}.

\itt{perturb\_to\_make\_definite} is a scalar variable of type default \logical,
that must be set \true\ if the user wants to guarantee that the
computed preconditioner is suitable by boosting the diagonal of the
requested $\bmG$ and \false\ otherwise.
The default is {\tt perturb\_to\_make\_definite = .TRUE.}.

\itt{get\_norm\_residual} is a scalar variable of type default \logical,
that must be set \true\ if the user wishes the package to return
the value of the norm of the residuals for the computed solution
when applying the preconditioner and  \false\ otherwise.
The default is {\tt get\_norm\_residual = .FALSE.}.

\itt{space\_critical} is a scalar variable of type default \logical,
that must be set \true\ if space is critical when allocating arrays
and  \false\ otherwise. The package may run faster if
{\tt space\_critical} is \false\ but at the possible expense of a larger
storage requirement. The default is {\tt space\_critical = .FALSE.}.

\itt{deallocate\_error\_fatal} is a scalar variable of type default \logical,
that must be set \true\ if the user wishes to terminate execution if
a deallocation  fails, and \false\ if an attempt to continue
will be made. The default is {\tt deallocate\_error\_fatal = .FALSE.}.

\itt{symmetric\_linear\_solver} is a scalar variable of type default \character\
and length 30, that specifies the external package to be used to
solve any general symmetric linear systems that might arise. Possible
choices are
{\tt 'sils'},
{\tt 'ma27'},
{\tt 'ma57'},
{\tt 'ma77'},
{\tt 'ma86'},
{\tt 'ma97'},
{\tt 'ssids'},
{\tt 'pardiso'}
and
{\tt 'wsmp'},
although only {\tt 'sils'} and, for OMP 4.0-compliant compilers,
{\tt 'ssids'} are installed by default.
See the documentation for the \galahad\ package {\tt SLS} for further details.
The default is {\tt symmetric\_linear\_solver = 'sils'}.

\itt{definite\_linear\_solver} is a scalar variable of type default \character\
and length 30, that specifies the external package to be used to
solve any symmetric positive-definite linear system that might arise. Possible
choices are
{\tt 'sils'},
{\tt 'ma27'},
{\tt 'ma57'},
{\tt 'ma77'},
{\tt 'ma86'},
{\tt 'ma87'},
{\tt 'ma97'},
{\tt 'ssids'},
{\tt 'pardiso'}
and
{\tt 'wsmp'},
although only {\tt 'sils'} and, for OMP 4.0-compliant compilers,
{\tt 'ssids'} are installed by default.
See the documentation for the \galahad\ package {\tt SLS} for further details.
The default is {\tt definite\_linear\_solver = 'sils'}.

\itt{unsymmetric\_linear\_solver} is a scalar variable of type default
\character\
and length 30, that specifies the external package to be used to
solve any unsymmetric linear systems that might arise. Possible
choices are
{\tt 'gls'},
{\tt 'ma28'}
and
{\tt 'ma48'},
although only {\tt 'gls'} is installed by default.
See the documentation for the \galahad\ package {\tt ULS} for further details.
The default is {\tt unsymmetric\_linear\_solver = 'gls'}.

\itt{prefix} is a scalar variable of type default \character\
and length 30, that may be used to provide a user-selected
character string to preface every line of printed output.
Specifically, each line of output will be prefaced by the string
{\tt prefix(2:LEN(TRIM(prefix))-1)},
thus ignoring the first and last non-null components of the
supplied string. If the user does not want to preface lines by such
a string, they may use the default {\tt prefix = ""}.

\itt{SLS\_control} is a scalar variable argument of type
{\tt SLS\_control\_type} that is used to pass control
options to external packages used to solve any symmetric
linear systems that might arise.
See the documentation for the \galahad\ package {\tt SLS} for further details.
In particular, default values are as for {\tt SLS}.

\itt{ULS\_control} is a scalar variable argument of type
{\tt ULS\_control\_type} that is used to pass control
options to external packages used to solve any unsymmetric
linear systems that might arise.
See the documentation for the \galahad\ package {\tt ULS} for further details.
In particular, default values are as for {\tt ULS}.

\end{description}

%%%%%%%%%%% time type %%%%%%%%%%%

\subsubsection{The derived data type for holding timing
 information}\label{typetime}
The derived data type
{\tt \packagename\_time\_type}
is used to hold elapsed CPU and system clock times for the various parts
of the calculation. The components of
{\tt \packagename\_time\_type}
are:
\begin{description}
\itt{total} is a scalar variable of type default \realdp, that gives
 the total CPU time (in seconds) spent in the package.

\itt{form} is a scalar variable of type default \realdp, that gives
 the CPU time spent forming the preconditioner.

\itt{factorize} is a scalar variable of type default \realdp, that gives
 the CPU time spent factorizing the preconditioner.

\itt{apply} is a scalar variable of type default \realdp, that gives
 the CPU time spent applying the preconditioner.

\itt{clock\_total} is a scalar variable of type default \realdp, that gives
 the total elapsed system clock time (in seconds) spent in the package.

\itt{clock\_form} is a scalar variable of type default \realdp, that gives
 the  elapsed system clock time spent forming the preconditioner.

\itt{clock\_factorize} is a scalar variable of type default \realdp, that gives
 the elapsed system clock time spent factorizing the preconditioner.

\itt{clock\_apply} is a scalar variable of type default \realdp, that gives
 the elapsed system clock time spent applying the preconditioner.

\end{description}

%%%%%%%%%%% inform type %%%%%%%%%%%

\subsubsection{The derived data type for holding informational
 parameters}\label{typeinform}
The derived data type
{\tt \packagename\_inform\_type}
is used to hold parameters that give information about the progress and needs
of the algorithm. The components of
{\tt \packagename\_inform\_type}
are:

\begin{description}

\itt{status} is a scalar variable of type default \integer, that gives the
exit status of the algorithm.
%See Sections~\ref{galerrors} and \ref{galinfo}
See \S\ref{galerrors}
for details.

\itt{alloc\_status} is a scalar variable of type default \integer, that gives
the status of the last attempted array allocation or deallocation.
This will be 0 if {\tt status = 0}.

\itt{bad\_alloc} is a scalar variable of type default \character\
and length 80, that  gives the name of the last internal array
for which there were allocation or deallocation errors.
This will be the null string if {\tt status = 0}.

%\itt{sls\_analyse\_status} is a scalar variable of type default \integer
%that reports the return code from the most recent call to {\tt SLS\_analyse}
%by {\tt \packagename\_form\_and\_factorize}. A non-zero value indicates
%a warning or an error. See the documentation for the package {\tt SLS}
%for further details.

%\itt{sls\_factorize\_status} is a scalar variable of type default \integer
%that reports the return code from the most recent call to {\tt SLS\_factorize}
%by {\tt \packagename\_form\_and\_factorize}. A non-zero value indicates
%a warning or an error. See the documentation for the package {\tt SLS}
%for further details.

%\itt{sls\_solve\_status} is a scalar variable of type default \integer
%that reports the return code from the most recent call to {\tt SLS\_solve}
%by {\tt \packagename\_solve}. A non-zero value indicates
%a warning or an error. See the documentation for the package {\tt SLS}
%for further details.

%\itt{uls\_factorize\_status} is a scalar variable of type default \integer,
%that reports the return code from the most recent call to {\tt ULS\_factorize}
%by {\tt \packagename\_form\_and\_factorize}. A non-zero value indicates
%a warning or an error. See the documentation for the package {\tt ULS}
%for further details.

%\itt{uls\_solve\_status} is a scalar variable of type default \integer,
%that reports the return code from the most recent call to {\tt ULS\_solve}
%by {\tt \packagename\_solve}. A non-zero value indicates
%a warning or an error. See the documentation for the package {\tt ULS}
%for further details.

\itt{factorization\_integer} is a scalar variable of type long \integer,
that reports the number of integers required to hold the factorization.

\itt{factorization\_real} is a scalar variable of type long \integer,
that reports the number of reals required to hold the factorization.

\itt{preconditioner} is a scalar variable of type default \integer,
that indicates the preconditioner method used. The range of values returned
corresponds to those requested in {\tt control\%preconditioner},
excepting that the requested value may have been altered to a more
appropriate one during the factorization. In particular, if the automatic
choice {\tt control\%preconditioner = 0} is requested,
{\tt preconditioner} reports the actual choice made.

\itt{factorization} is a scalar variable of type default \integer,
that indicates the factorization method used. The range of values returned
corresponds to those requested in {\tt control\%factorization},
excepting that the requested value may have been altered to a more
appropriate one during the factorization.
In particular, if the automatic
choice {\tt control\%factorization = 0} is requested,
{\tt factorization} reports the actual choice made.

\itt{rank} is a scalar variable of type default \integer, that gives the
computed rank of $\bmA$.

\itt{rank\_def} is a scalar variable of type default \logical, that has the
value \true\ if {\tt \packagename\_form\_and\_factorize} believes that
$\bmA$ is rank defficient, and \false\ otherwise

\itt{perturbed} is a scalar variable of type default \logical, that has the
value \true\ if and only if the original choice of $\bmG$ has been perturbed to
ensure that $\bmK_G$ is an appropriate preconditioner. This will only
happen if {\tt control\-\%perturb\_to\_make\_definite} has been set \true.

\itt{norm\_residual} is a scalar variable of type default \realdp,
that holds the infinity norm of the residual of the system \req{ls}
after a call to {\tt \packagename\_solve} if {\tt control\%get\_norm\_residual}
has been set \true. Otherwise it will have the value -1.0.

\ittf{time} is a scalar variable of type {\tt \packagename\_time\_type}
whose components are used to hold elapsed CPU and system clock times
(in seconds) for the various parts of the calculation
(see Section~\ref{typetime}).

\itt{SLS\_inform} is a scalar variable argument of type
{\tt SLS\_inform\_type} that is used to pass information
concerning the progress of the external packages used to solve any symmetric
linear systems that might arise.
See the documentation for the \galahad\ package {\tt SLS} for further details.

\itt{ULS\_inform} is a scalar variable argument of type
{\tt ULS\_inform\_type} that is used to pass information
concerning the progress of the external packages used to solve any unsymmetric
linear systems that might arise.
See the documentation for the \galahad\ package {\tt ULS} for further details.
\end{description}

%%%%%%%%%%% data type %%%%%%%%%%%

\subsubsection{The derived data type for holding problem data}\label{typedata}
The derived data types
{\tt \packagename\_explicit\_factor\_type},
{\tt \packagename\_implicit\_factor\_type} and
{\tt \packagename\_data\_type}
are used to hold all the data for the problem and the factors of
its preconditioners between calls of
{\tt \packagename} procedures.
This data should be preserved, untouched, from the initial call to
{\tt \packagename\_initialize}
to the final call to
{\tt \packagename\_terminate}.

%%%%%%%%%%%%%%%%%%%%%% argument lists %%%%%%%%%%%%%%%%%%%%%%%%

\galarguments
There are four procedures for user calls
(see \S\ref{galfeatures} for further features):

\begin{enumerate}
\item The subroutine
      {\tt \packagename\_initialize}
      is used to set default values, and initialize private data,
      before solving one or more problems with the
      same sparsity and bound structure.
\item The subroutine
      {\tt \packagename\_form\_and\_factorize}
      is called to form and factorize the preconditioner.
\item The subroutine
      {\tt \packagename\_solve}
      is called to apply the preconditioner, that is to solve a linear
      system of the form \req{ls}.
\item The subroutine
      {\tt \packagename\_terminate}
      is provided to allow the user to automatically deallocate array
       components of the private data, allocated by
       {\tt \packagename\_form\_and\_factorize}
       at the end of the solution process.
\end{enumerate}
We use square brackets {\tt [ ]} to indicate \optional arguments.

%%%%%% initialization subroutine %%%%%%

\subsubsection{The initialization subroutine}\label{subinit}
 Default values are provided as follows:
\vspace*{1mm}

\hspace{8mm}
{\tt CALL \packagename\_initialize( data, control, inform )}

\vspace*{-3mm}
\begin{description}

\itt{data} is a scalar \intentinout\ argument of type
{\tt \packagename\_data\_type}
(see \S\ref{typedata}). It is used to hold data about the problem being
solved.

\itt{control} is a scalar \intentout\ argument of type
{\tt \packagename\_control\_type}
(see \S\ref{typecontrol}).
On exit, {\tt control} contains default values for the components as
described in \S\ref{typecontrol}.
These values should only be changed after calling
{\tt \packagename\_initialize}.

\itt{inform} is a scalar \intentout\ argument of type
{\tt \packagename\_inform\_type}
(see Section~\ref{typeinform}). A successful call to
{\tt \packagename\_initialize}
is indicated when the  component {\tt status} has the value 0.
For other return values of {\tt status}, see Section~\ref{galerrors}.

\end{description}

%%%%%%%%% main solution subroutine %%%%%%

\subsubsection{The subroutine for forming and factorizing the preconditioner}
A preconditioner of the form \req{prec1} is formed and factorized as follows:
\vspace*{1mm}

\hspace{8mm}
{\tt CALL \packagename\_form\_and\_factorize( n, m, H, A, C, data, control, inform[, D] )}

%\vspace*{-3mm}
\begin{description}
\itt{n} is a scalar \intentin\ argument of type default \integer\ that specifies
the number of rows of $\bmH$ (and columns of $\bmA$).

\itt{m} is a scalar \intentin\ argument of type default \integer\ that specifies
the number of rows of $\bmA$ and $\bmC$.

\itt{H} is a scalar \intentin\ argument of type {\tt SMT\_type} whose
components must be set to specify the data defining the matrix $\bmH$
(see \S\ref{typesmt}).

\itt{A} is a scalar \intentin\ argument of type {\tt SMT\_type} whose
components must be set to specify the data defining the matrix $\bmA$
(see \S\ref{typesmt}).

\itt{C} is a scalar \intentin\ argument of type {\tt SMT\_type} whose
components must be set to specify the data defining the matrix $\bmC$
(see \S\ref{typesmt}).

\itt{data} is a scalar \intentinout\ argument of type
{\tt \packagename\_data\_type}
(see \S\ref{typedata}). It is used to hold data about the problem being
solved. It must not have been altered {\bf by the user} since the last call to
{\tt \packagename\_initialize}.

\itt{control} is a scalar \intentin\ argument of type
{\tt \packagename\_control\_type}
(see \S\ref{typecontrol}). Default values may be assigned by calling
{\tt \packagename\_initialize} prior to the first call to
{\tt \packagename\_solve}.

\itt{inform} is a scalar \intentout\ argument of type
{\tt \packagename\_inform\_type}
(see \S\ref{typeinform}). A successful call to
{\tt \packagename\_solve}
is indicated when the  component {\tt status} has the value 0.
For other return values of {\tt status}, see \S\ref{galerrors}.

\itt{D} is a rank-one \optional\ \intentin\ argument of type \realdp\
and length at least {\tt n}, whose $i$-th component give the value of the
$i$-th diagonal entry of the matrix $\bmD$ used when
{\tt control\%pre\-conditioner = 5} (see \S\ref{typecontrol}).
{\tt D} need not be \present\ for other values of {\tt control\%preconditio\-ner}.

\end{description}

\subsubsection{The subroutine for applying the preconditioner}
The preconditioner may be applied to solve a system of the
form \req{ls} as follows:
\vspace*{1mm}

\hspace{8mm}
{\tt CALL \packagename\_solve(n, m, A, C, data, control, inform, SOL )}
\vspace*{1mm}

\noindent
Components {\tt n}, {\tt m},  {\tt A}, {\tt C}, {\tt data} and
{\tt control} are exactly as described for
{\tt \packagename\_form\_and\_factorize} and must not have been
altered in the interim.

\vspace*{-3mm}
\begin{description}

\itt{inform} is a scalar \intentout\ argument of type
{\tt \packagename\_inform\_type}
(see \S\ref{typeinform}), that should be passed unaltered since
the last call to {\tt \packagename\_form\_and\_factorize} or
{\tt \packagename\_solve}.  A successful call to
{\tt \packagename\_solve}
is indicated when the  component {\tt status} has the value 0.
For other return values of {\tt status}, see \S\ref{galerrors}.

\itt{SOL} is a rank-one  \intentinout\ array of type default \real\
and length at least {\tt n+m}, that must be set on entry to hold
the composite vector $( a^T \;\; b^T)^T$.
In particular {\tt SOL(}$i${\tt )}, $i = 1,$ \ldots {\tt n} should be
set to $a_i$, and
{\tt SOL(n}$+j${\tt )}, $j = 1, \ldots,$ {\tt m} should be
set to $b_j$. On successful exit, {\tt SOL}
will contain the solution $( x^T \;\; y^T)^T$ to \req{ls}, that is
{\tt SOL(}$i${\tt )}, $i = 1, \ldots,$ {\tt n} will give
$x_i$, and  {\tt SOL(n}$+j${\tt )}, $j = 1,\ldots,$ {\tt m} will contain
$y_j$.



\end{description}

%%%%%%% termination subroutine %%%%%%

\subsubsection{The  termination subroutine}
All previously allocated arrays are deallocated as follows:
\vspace*{1mm}

\hspace{8mm}
{\tt CALL \packagename\_terminate( data, control, inform )}

%\vspace*{-3mm}
\begin{description}

\itt{data} is a scalar \intentinout\ argument of type
{\tt \packagename\_data\_type}
exactly as for
{\tt \packagename\_solve},
which must not have been altered {\bf by the user} since the last call to
{\tt \packagename\_initialize}.
On exit, array components will have been deallocated.

\itt{control} is a scalar \intentin\ argument of type
{\tt \packagename\_control\_type}
exactly as for
{\tt \packagename\_solve}.

\itt{inform} is a scalar \intentout\ argument of type
{\tt \packagename\_inform\_type}
exactly as for
{\tt \packagename\_solve}.
Only the component {\tt status} will be set on exit, and a
successful call to
{\tt \packagename\_terminate}
is indicated when this  component {\tt status} has the value 0.
For other return values of {\tt status}, see \S\ref{galerrors}.

\end{description}

%%%%%%%%%%%%%%%%%%%%%% Warning and error messages %%%%%%%%%%%%%%%%%%%%%%%%

\galerrors
A negative value of {\tt inform\%status} on exit from
{\tt \packagename\_form\_and\_factorize},
{\tt \packagename\_solve}
or
{\tt \packagename\_terminate}
indicates that an error has occurred. No further calls should be made
until the error has been corrected. Possible values are:

\begin{description}

\itt{\galerrallocate.} An allocation error occurred.
A message indicating the offending
array is written on unit {\tt control\%error}, and the returned allocation
status and a string containing the name of the offending array
are held in {\tt inform\%alloc\_\-status}
and {\tt inform\%bad\_alloc} respectively.

\itt{\galerrdeallocate.} A deallocation error occurred.
A message indicating the offending
array is written on unit {\tt control\%error} and the returned allocation
status and a string containing the name of the offending array
are held in {\tt inform\%alloc\_\-status}
and {\tt inform\%bad\_alloc} respectively.

\itt{\galerrrestrictions.} One of the restrictions
   {\tt prob\%n} $> 0$ or {\tt prob\%m} $\geq  0$
    or requirements that {\tt prob\%A\_type}, {\tt prob\%H\_type} and
    {\tt prob\%C\_type} contain its relevant string
    {\tt 'DENSE'}, {\tt 'COORDINATE'}, {\tt 'SPARSE\_BY\_ROWS'},
    {\tt 'DIAGONAL'}, {\tt 'SCALED\_IDENTITY'}, {\tt 'IDENTITY'},
    {\tt ZERO} or {\tt NONE}
    has been violated.

%\itt{-3.} At least one of the arrays
% {\tt p\%A\_val}, {\tt p\%A\_row}, {\tt p\%A\_col},
% {\tt p\%H\_val}, {\tt p\%H\_row} or {\tt p\%H\_col},
% is not large enough to hold the original, or reordered, matrices $\bmA$
% or $\bmH$.

\itt{\galerranalysis.} An error was reported by {\tt SLS\_analyse}. The return
status from {\tt SLS\_analyse} is given in
{\tt inform\%SLS\_inform\%status}.
See the documentation for the \galahad\ package {\tt SLS} for further details.

\itt{\galerrfactorization.} An error was reported by {\tt SLS\_factorize}.
The return status from {\tt SLS\_factorize} is given in
{\tt inform\%SLS\_inform\%status}.
See the documentation for the \galahad\ package {\tt SLS} for further details.

\itt{\galerrsolve.} An error was reported by {\tt SLS\_solve}. The return
status from {\tt SLS\_solve} is given in {\tt inform\%SLS\_inform\%status}.
See the documentation for the \galahad\ package {\tt SLS} for further details.

\itt{\galerrulsfactorization.} An error was reported by {\tt ULS\_factorize}.
The return status from {\tt ULS\_factorize} is given in
{\tt inform\%uls\_factorize\_\-status}.
See the documentation for the \galahad\ package {\tt ULS} for further details.

\itt{\galerrulssolve.} An error was reported by {\tt ULS\_solve}. The return
status from {\tt ULS\_solve} is given in {\tt inform\%uls\_solve\_status}.
See the documentation for the \galahad\ package {\tt ULS} for further details.

\itt{\galerrpreconditioner.} The computed preconditioner is singular
and is thus unsuitable.

\itt{\galerrinertia.} The computed preconditioner has the wrong
inertia and is thus unsuitable.

\itt{\galerrsort.} An error was reported by {\tt SORT\_reorder\_by\_rows}.
The return status from {\tt SORT\_reorder\_by\_rows}
is given in {\tt inform\%sort\_status}.
See the documentation for the \galahad\ package {\tt SORT} for further details.

\itt{\galunknownsolver.} The requested linear equation solver is not available.

\end{description}

A positive value of {\tt inform\%status} on exit from
{\tt \packagename\_form\_and\_factorize} warns of unexpected behaviour.
A possible values is:

\begin{description}

\itt{1.} The matrx $\bmA$ is rank defficient.

\end{description}

%%%%%%%%%%%%%%%%%%%%%% Further features %%%%%%%%%%%%%%%%%%%%%%%%

\galfeatures
\noindent In this section, we describe an alternative means of setting
control parameters, that is components of the variable {\tt control} of type
{\tt \packagename\_control\_type}
(see \S\ref{typecontrol}),
by reading an appropriate data specification file using the
subroutine {\tt \packagename\_read\_specfile}. This facility
is useful as it allows a user to change  {\tt \packagename} control parameters
without editing and recompiling programs that call {\tt \packagename}.

A specification file, or specfile, is a data file containing a number of
"specification commands". Each command occurs on a separate line,
and comprises a "keyword",
which is a string (in a close-to-natural language) used to identify a
control parameter, and
an (optional) "value", which defines the value to be assigned to the given
control parameter. All keywords and values are case insensitive,
keywords may be preceded by one or more blanks but
values must not contain blanks, and
each value must be separated from its keyword by at least one blank.
Values must not contain more than 30 characters, and
each line of the specfile is limited to 80 characters,
including the blanks separating keyword and value.



The portion of the specification file used by
{\tt \packagename\_read\_specfile}
must start
with a "{\tt BEGIN \packagename}" command and end with an
"{\tt END}" command.  The syntax of the specfile is thus defined as follows:
\begin{verbatim}
  ( .. lines ignored by SBLS_read_specfile .. )
    BEGIN SBLS
       keyword    value
       .......    .....
       keyword    value
    END
  ( .. lines ignored by SBLS_read_specfile .. )
\end{verbatim}
where keyword and value are two strings separated by (at least) one blank.
The ``{\tt BEGIN \packagename}'' and ``{\tt END}'' delimiter command lines
may contain additional (trailing) strings so long as such strings are
separated by one or more blanks, so that lines such as
\begin{verbatim}
    BEGIN SBLS SPECIFICATION
\end{verbatim}
and
\begin{verbatim}
    END SBLS SPECIFICATION
\end{verbatim}
are acceptable. Furthermore,
between the
``{\tt BEGIN \packagename}'' and ``{\tt END}'' delimiters,
specification commands may occur in any order.  Blank lines and
lines whose first non-blank character is {\tt !} or {\tt *} are ignored.
The content
of a line after a {\tt !} or {\tt *} character is also
ignored (as is the {\tt !} or {\tt *}
character itself). This provides an easy manner to "comment out" some
specification commands, or to comment specific values
of certain control parameters.

The value of a control parameters may be of three different types, namely
integer, logical or real.
Integer and real values may be expressed in any relevant Fortran integer and
floating-point formats (respectively). Permitted values for logical
parameters are "{\tt ON}", "{\tt TRUE}", "{\tt .TRUE.}", "{\tt T}",
"{\tt YES}", "{\tt Y}", or "{\tt OFF}", "{\tt NO}",
"{\tt N}", "{\tt FALSE}", "{\tt .FALSE.}" and "{\tt F}".
Empty values are also allowed for
logical control parameters, and are interpreted as "{\tt TRUE}".

The specification file must be open for
input when {\tt \packagename\_read\_specfile}
is called, and the associated device number
passed to the routine in device (see below).
Note that the corresponding
file is {\tt REWIND}ed, which makes it possible to combine the specifications
for more than one program/routine.  For the same reason, the file is not
closed by {\tt \packagename\_read\_specfile}.

Control parameters corresponding to the components
{\tt SLS\_control}
and
{\tt ULS\_control} may be changed by including additional sections enclosed by
``{\tt BEGIN SLS}'' and
``{\tt END SLS}'', and
``{\tt BEGIN ULS}'' and
``{\tt END ULS}'', respectively.
See the specification sheets for the packages
{\tt \libraryname\_SLS}
and
{\tt \libraryname\_ULS}
for further details.

\subsubsection{To read control parameters from a specification file}
\label{readspec}

Control parameters may be read from a file as follows:
\hskip0.5in

\def\baselinestretch{0.8}
{\tt
\begin{verbatim}
     CALL SBLS_read_specfile( control, device )
\end{verbatim}
}
\def\baselinestretch{1.0}

\begin{description}
\itt{control} is a scalar \intentinout argument of type
{\tt \packagename\_control\_type}
(see \S\ref{typecontrol}).
Default values should have already been set, perhaps by calling
{\tt \packagename\_initialize}.
On exit, individual components of {\tt control} may have been changed
according to the commands found in the specfile. Specfile commands and
the component (see \S\ref{typecontrol}) of {\tt control}
that each affects are given in Table~\ref{specfile}.
\bctable{|l|l|l|}
\hline
  command & component of {\tt control} & value type \\
\hline

  {\tt error-printout-device} & {\tt \%error} & integer \\
  {\tt printout-device} & {\tt \%out} & integer \\
  {\tt print-level} & {\tt \%print\_level} & integer \\
%  {\tt initial-workspace-for-unsymmetric-solver} & {\tt \%len\_ulsmin} & integer \\
%  {\tt initial-integer-workspace}  & {\tt \%indmin} & integer \\
%  {\tt initial-real-workspace}  & {\tt \%valmin} & integer \\
  {\tt maximum-refinements}  & {\tt \%itref\_max} & integer \\
  {\tt preconditioner-used} & {\tt \%preconditioner} & integer \\
  {\tt semi-bandwidth-for-band-preconditioner} & {\tt \%semi\_bandwidth} & integer \\
  {\tt factorization-used} & {\tt \%factorization} & integer \\
  {\tt maximum-column-nonzeros-in-schur-complement}  & {\tt \%max\_col} & integer \\
%  {\tt ordering-used}  & {\tt \%ordering} & integer \\
%  {\tt scaling-used}   & {\tt \%scaling} & integer \\
  {\tt has-a-changed}   & {\tt \%new\_a} & integer \\
  {\tt has-h-changed}  & {\tt \%new\_h} & integer \\
  {\tt has-c-changed}  & {\tt \%new\_c} & integer \\
  {\tt minimum-diagonal}  & {\tt \%min\_diagonal} & real \\
%  {\tt pivot-tolerance-used}  & {\tt \%pivot\_tol} & real \\
  {\tt pivot-tolerance-used-for-basis}  & {\tt \%pivot\_tol\_for\_basis} & real \\
%  {\tt zero-pivot-tolerance}  & {\tt \%zero\_pivot} & real \\
%  {\tt static-pivoting-diagonal-perturbation}  & {\tt \%static\_tolerance} & real \\
%  {\tt level-at-which-to-switch-to-static}  & {\tt \%static\_level} & real \\
  {\tt find-basis-by-transpose}  & {\tt \%find\_basis\_by\_transpose} & logical \\
  {\tt remove-linear-dependencies}  & {\tt \%remove\_dependencies} & logical \\
  {\tt affine-constraints}  & {\tt \%affine} & logical \\
  {\tt check-for-reliable-basis}  & {\tt \%check\_basis} & logical \\
  {\tt perturb-to-make-+ve-definite}   & {\tt \%perturb\_to\_make\_definite} & logical \\
  {\tt get-norm-residual}   & {\tt \%get\_norm\_residual} & logical \\
  {\tt space-critical}   & {\tt \%space\_critical} & logical \\
  {\tt deallocate-error-fatal}   & {\tt \%deallocate\_error\_fatal} & logical \\
  {\tt symmetric-linear-equation-solver} & {\tt \%symmetric\_linear\_solver} & character \\
  {\tt definite-linear-equation-solver} & {\tt \%definite\_linear\_solver} & character \\
  {\tt unsymmetric-linear-equation-solver} & {\tt \%unsymmetric\_linear\_solver} & character \\
  {\tt output-line-prefix} & {\tt \%prefix} & character \\
\hline

\ectable{\label{specfile}Specfile commands and associated
components of {\tt control}.}

%Individual components of the components {\tt control\%SLS\_control}
%and {\tt control\%ULS\_control}
%may be changed by specifying separate specfiles of the form
%\begin{verbatim}
%    BEGIN SLS
%     .......
%    END
%\end{verbatim}
%and
%\begin{verbatim}
%    BEGIN ULS
%     .......
%%    END
%\end{verbatim}
%just as described in the documentation for the \galahad\ packages
%{\tt SLS} and {\tt ULS}.

\itt{device} is a scalar \intentin argument of type default \integer,
that must be set to the unit number on which the specfile
has been opened. If {\tt device} is not open, {\tt control} will
not be altered and execution will continue, but an error message
will be printed on unit {\tt control\%error}.

\end{description}

%%%%%%%%%%%%%%%%%%%%%% Information printed %%%%%%%%%%%%%%%%%%%%%%%%

\galinfo
If {\tt control\%print\_level} is positive, information about the progress
of the algorithm will be printed on unit {\tt control\-\%out}.
If {\tt control\%print\_level} $= 1$, statistics concerning the factorization,
as well as warning and error messages will be reported.
If {\tt control\%print\_level} $= 2$, additional information about the
progress of the factorization and the solution phases will be given.
If {\tt control\%print\_level} $> 2$, debug information, of little
interest to the general user, may be returned.

%%%%%%%%%%%%%%%%%%%%%% GENERAL INFORMATION %%%%%%%%%%%%%%%%%%%%%%%%

\galgeneral

\galcommon None.
\galworkspace Provided automatically by the module.
\galroutines None.
\galmodules {\tt \packagename\_solve} calls the \galahad\ packages
{\tt GALAHAD\_CLOCK},
{\tt GALAHAD\_SY\-M\-BOLS}, \\
{\tt GALAHAD\-\_SPACE},
{\tt GALAHAD\_SMT},
{\tt GALAHAD\_QPT},
{\tt GALAHAD\_SLS},
{\tt GALAHAD\_ULS} and
{\tt GALAHAD\_SPECFILE},
\galio Output is under control of the arguments
 {\tt control\%error}, {\tt control\%out} and {\tt control\%print\_level}.
\galrestrictions {\tt prob\%n} $> 0$, {\tt prob\%m} $\geq  0$,
{\tt prob\%H\_type}, {\tt prob\%A\_type}
and {\tt prob\%C\_type} $\in \{${\tt 'DENSE'},
 {\tt 'COORDINATE'}, {\tt 'SPARSE\_BY\_ROWS'}, {\tt 'DIAGONAL'},
 {\tt 'SCALED\_IDENTITY'}, {\tt 'IDENTITY'} $\}$,
 {\tt ZERO}, {\tt NONE}.
\galportability ISO Fortran~95 + TR 15581 or Fortran~2003.
The package is thread-safe.

%%%%%%%%%%%%%%%%%%%%%% METHOD %%%%%%%%%%%%%%%%%%%%%%%%

\galmethod
The method used depends on whether an explicit or implicit
factorization is required. In the explicit case, the
package is really little more than a wrapper for the
symmetric, indefinite linear solver {\tt SLS} in
which the system matrix $\bmK_G$ is assembled from its constituents
$\bmA$, $\bmC$ and whichever $\bmG$ is requested by the user.
Implicit-factorization preconditioners are more involved,
and there is a large variety of different possibilities. The
essential ideas are described in detail in
\vspace*{1mm}

\noindent
H. S. Dollar, N. I. M. Gould and A. J. Wathen.
``On implicit-factorization constraint preconditioners''.
In  Large Scale Nonlinear Optimization (G. Di Pillo and M. Roma, eds.)
Springer Series on Nonconvex Optimization and Its Applications, Vol. 83,
Springer Verlag (2006) 61--82

\noindent
and

\noindent
H. S. Dollar, N. I. M. Gould, W. H. A. Schilders and A. J. Wathen
``On iterative methods and implicit-factorization preconditioners for
regularized saddle-point systems''.
SIAM Journal on Matrix Analysis and Applications, {\bf 28(1)} (2006) 170--189.
\vspace*{1mm}

\noindent
The range-space factorization is based upon the decomposition
\disp{ \bmK_{G} =
\mat{cc}{ \bmG & \bmzero \\ \bmA & \bmI}
\mat{cc}{ \bmG^{-1} & \bmzero \\ \bmzero & - \bmS}
\mat{cc}{ \bmG & \bmA^T \\ \bmzero & \bmI},}
where the ``Schur complement'' $\bmS = \bmC + \bmA \bmG^{-1} \bmA^T$.
Such a method requires that $\bmS$ is easily invertible. This is often the
case when $\bmG$ is a diagonal matrix, in which case $\bmS$ is frequently
sparse, or when $m \ll n$ in which case $\bmS$
is small and a dense Cholesky factorization may be used.
\vspace*{1mm}

\noindent
When $\bmC = 0$, the null-space factorization is based upon the decomposition
\disp{ \bmK_{G} = \bmP
\mat{ccc}{
\bmG_{11}^{} & \bmzero & \bmI \\
\bmG_{21}^{} & \bmI & \bmA_{2}^{T} \bmA_{1}^{-T}  \\
\bmA_{1}^{} & \bmzero & \bmzero }
\mat{ccc}{\bmzero & \bmzero & \bmI \\
\;\;\; \bmzero \;\; & \;\; \bmR \;\; & \bmzero \\
\bmI & \bmzero & - \bmG_{11}^{}}
\mat{ccc}{
\bmG_{11}^{} & \bmG_{21}^T & \bmA_{1}^T \\
\bmzero & \bmI & \bmzero \\
\bmI & \bmA_{1}^{-1} \bmA_{2}^{} & \bmzero}
\bmP^T,}
where the ``reduced Hessian''
\disp{\bmR =
( - \bmA_{2}^{T} \bmA_1^{-T} \;\; \bmI )
\mat{cc}{\bmG_{11}^{} & \bmG_{21}^{T} \\ \bmG_{21}^{} & \bmG_{22}^{}}
\vect{ - \bmA_1^{-1} \bmA_2^{} \\ \bmI}}
and $\bmP$ is a suitably-chosen permutation for which $\bmA_1$ is invertible.
The method is most useful when $m \approx n$ as then the dimension
of $\bmR$ is small and a dense Cholesky factorization may be used.


%%%%%%%%%%%%%%%%%%%%%% EXAMPLE %%%%%%%%%%%%%%%%%%%%%%%%

\galexample
Suppose we wish to solve the linear system \req{ls} with matrix data
\disp{\bmH = \mat{ccc}{1 & & 4 \\ & 2 & \\ 4 &  & 3}, \;\;
 \bmA = \mat{ccc}{ 2 & 1 & \\ & 1 & 1}
\tim{and}  \bmC = \mat{cc}{  & 1 \\ 1 &   }}
and right-hand sides
\disp{\bma = \vect{7 \\ 4 \\ 8} \tim{and} \bmb = \vect{2 \\ 1}.}
Then storing the matrices in sparse co-ordinate format,
we may use the following code:

{\tt \small
\VerbatimInput{\packageexample}
}
\noindent
This produces the following output:
{\tt \small
\VerbatimInput{\packageresults}
}
\noindent
The same problem may be solved holding the data in
a sparse row-wise storage format by replacing the lines
{\tt \small
\begin{verbatim}
!  sparse co-ordinate storage format
...
! problem data complete
\end{verbatim}
}
\noindent
by
{\tt \small
\begin{verbatim}
! sparse row-wise storage format
   CALL SMT_put( H%type, 'SPARSE_BY_ROWS', s )  ! Specify sparse-by-rows
   CALL SMT_put( A%type, 'SPARSE_BY_ROWS', s )  ! storage for H, A and C
   CALL SMT_put( C%type, 'SPARSE_BY_ROWS', s )
   ALLOCATE( H%val( h_ne ), H%col( h_ne ), H%ptr( n + 1 ) )
   ALLOCATE( A%val( a_ne ), A%col( a_ne ), A%ptr( m + 1 ) )
   ALLOCATE( C%val( c_ne ), C%col( c_ne ), C%ptr( m + 1 ) )
   H%val = (/ 1.0_wp, 2.0_wp, 3.0_wp, 4.0_wp /) ! matrix H
   H%col = (/ 1, 2, 3, 1 /)                     ! NB lower triangular
   H%ptr = (/ 1, 2, 3, 5 /)                     ! Set row pointers
   A%val = (/ 2.0_wp, 1.0_wp, 1.0_wp, 1.0_wp /) ! matrix A
   A%col = (/ 1, 2, 2, 3 /)
   A%ptr = (/ 1, 3, 5 /)                        ! Set row pointers
   C%val = (/ 1.0_wp /)                         ! matrix C
   C%col = (/ 1 /)                              ! NB lower triangular
   C%ptr = (/ 1, 1, 2 /)                        ! Set row pointers
! problem data complete
\end{verbatim}
}
\noindent
or using a dense storage format with the replacement lines
{\tt \small
\begin{verbatim}
! dense storage format
   CALL SMT_put( H%type, 'DENSE', s )  ! Specify dense
   CALL SMT_put( A%type, 'DENSE', s )  ! storage for H, A and C
   CALL SMT_put( C%type, 'DENSE', s )
   ALLOCATE( H%val( n * ( n + 1 ) / 2 ) )
   ALLOCATE( A%val( n * m ) )
   ALLOCATE( C%val( m * ( m + 1 ) / 2 ) )
   H%val = (/ 1.0_wp, 0.0_wp, 2.0_wp, 4.0_wp, 0.0_wp, 3.0_wp /) ! H
   A%val = (/ 2.0_wp, 1.0_wp, 0.0_wp, 0.0_wp, 1.0_wp, 1.0_wp /) ! A
   C%val = (/ 0.0_wp, 1.0_wp, 0.0_wp /)                         ! C
! problem data complete
\end{verbatim}
}
\noindent
respectively.

If instead $\bmH$ had been the diagonal matrix
\disp{\bmH = \mat{ccc}{1 & &   \\ & 0 & \\  &  & 3}}
but the other data is as before, the diagonal storage scheme
might be used for $\bmH$, and in this case we would instead
{\tt \small
\begin{verbatim}
   CALL SMT_put( prob%H%type, 'DIAGONAL', s )  ! Specify dense storage for H
   ALLOCATE( p%H%val( n ) )
   p%H%val = (/ 1.0_wp, 0.0_wp, 3.0_wp /) ! Hessian values
\end{verbatim}
}
\noindent
Notice here that zero diagonal entries are stored.

\end{document}
