\documentclass{galahad}

% set the release and package names

\newcommand{\package}{ssls}
\newcommand{\packagename}{SSLS}
\newcommand{\fullpackagename}{\libraryname\_\packagename}

\begin{document}

\galheader

%%%%%%%%%%%%%%%%%%%%%% SUMMARY %%%%%%%%%%%%%%%%%%%%%%%%

\galsummary
Given a matrix $\bmA$ and symmetric matrices $\bmH$ and $\bmC$,
{\bf form and factorize the block, symmetric matrix}
\disp{ \bmK = \mat{cc}{ \bmH & \bmA^T \\ \bmA  & - \bmC },}
and subsequently {\bf solve systems}
\eqn{ls}{ \mat{cc}{ \bmH & \bmA^T \\ \bmA  & - \bmC } \vect{ \bmx \\ \bmy }
 = \vect{\bma \\ \bmb},}
using the \galahad\ symmetric-indefinite factorization package {\tt SLS}.
Full advantage is taken of any zero coefficients in the matrices $\bmH$,
$\bmA$ and $\bmC$.

%%%%%%%%%%%%%%%%%%%%%% attributes %%%%%%%%%%%%%%%%%%%%%%%%

\galattributes
\galversions{\tt  \fullpackagename\_single, \fullpackagename\_double}.
\galuses {\tt GALAHAD\_CLOCK},
{\tt GALAHAD\_\-SY\-M\-BOLS},
{\tt GALAHAD\-\_SPACE},
{\tt GALAHAD\_SMT},
{\tt GALAHAD\_QPT},
{\tt GALAHAD\_SLS},
{\tt GALAHAD\_SPECFILE},
\galdate July 2025.
\galorigin N. I. M. Gould,
Rutherford Appleton Laboratory.
\gallanguage Fortran~95 + TR 15581 or Fortran~2003.
\galparallelism Some options may use OpenMP and its runtime library.

%%%%%%%%%%%%%%%%%%%%%% HOW TO USE %%%%%%%%%%%%%%%%%%%%%%%%

\galhowto

\input{versions}

\noindent
If it is required to use more than one of the modules at the same time, 
the derived types
{\tt SMT\_type},
{\tt QPT\_problem\_type},
{\tt \packagename\_time\_type},
{\tt \packagename\_control\_type},
{\tt \packagename\_inform\_type}
and
{\tt \packagename\_data\_type}
(\S\ref{galtypes})
and the subroutines
{\tt \packagename\_\-initialize},
{\tt \packagename\_\-analyse},
{\tt \packagename\_\-factorize},
{\tt \packagename\_\-solve},
{\tt \packagename\_terminate},
(\S\ref{galarguments})
and
{\tt \packagename\_read\_specfile}
(\S\ref{galfeatures})
must be renamed on one of the {\tt USE} statements.

%%%%%%%%%%%%%%%%%%%%%% matrix formats %%%%%%%%%%%%%%%%%%%%%%%%

\galmatrix
Each of the input matrices $\bmH$, $\bmA$ and $\bmC$
may be stored in a variety of input formats.

\subsubsection{Dense storage format}\label{dense}
The matrix $\bmA$ is stored as a compact
dense matrix by rows, that is, the values of the entries of each row in turn are
stored in order within an appropriate real one-dimensional array.
Component $n \ast (i-1) + j$ of the storage array {\tt A\%val} will hold the
value $a_{ij}$ for $i = 1, \ldots , m$, $j = 1, \ldots , n$.
Since $\bmH$ and $\bmC$ are symmetric, only the lower triangular parts
(that is the part $h_{ij}$ for $1 \leq j \leq i \leq n$ and
$c_{ij}$ for $1 \leq j \leq i \leq m$) need be held. In these cases
the lower triangle will be stored by rows, that is
component $i \ast (i-1)/2 + j$ of the storage array {\tt H\%val}
will hold the value $h_{ij}$ (and, by symmetry, $h_{ji}$)
for $1 \leq j \leq i \leq n$. Similarly
component $i \ast (i-1)/2 + j$ of the storage array {\tt C\%val}
will hold the value $c_{ij}$ (and, by symmetry, $c_{ji}$)
for $1 \leq j \leq i \leq m$.

\subsubsection{Sparse co-ordinate storage format}\label{coordinate}
Only the nonzero entries of the matrices are stored. For the
$l$-th entry of $\bmA$, its row index $i$, column index $j$
and value $a_{ij}$
are stored in the $l$-th components of the integer arrays {\tt A\%row},
{\tt A\%col} and real array {\tt A\%val}, respectively.
The order is unimportant, but the total
number of entries {\tt A\%ne} is also required.
The same scheme is applicable to $\bmH$ and $\bmC$
(thus, for $\bmH$, requiring integer arrays {\tt H\%row}, {\tt H\%col}, a real
array  {\tt H\%val} and an integer value {\tt H\%ne}),
except that only the entries in the lower triangle need be stored.

\subsubsection{Sparse row-wise storage format}\label{rowwise}
Again only the nonzero entries are stored, but this time
they are ordered so that those in row $i$ appear directly before those
in row $i+1$. For the $i$-th row of $\bmA$, the $i$-th component of a
integer array {\tt A\%ptr} holds the position of the first entry in this row,
while {\tt A\%ptr} $(m+1)$ holds the total number of entries plus one.
The column indices $j$ and values $a_{ij}$ of the entries in the $i$-th row
are stored in components
$l =$ {\tt A\%ptr}$(i)$, \ldots ,{\tt A\%ptr} $(i+1)-1$ of the
integer array {\tt A\%col}, and real array {\tt A\%val}, respectively.
The same scheme is applicable to
$\bmH$ and $\bmC$ (thus, for $\bmH$,
requiring integer arrays {\tt H\%ptr}, {\tt H\%col}, and
a real array {\tt H\%val}),
except that only the entries in the lower triangle need be stored.

For sparse matrices, this scheme almost always requires less storage than
its predecessor.

\subsubsection{Diagonal storage format}\label{diagonal}
If $\bmH$ is diagonal (i.e., $h_{ij} = 0$ for all $1 \leq i \neq j \leq n$)
only the diagonals entries $h_{ii}$, $1 \leq i \leq n$,  need be stored,
and the first $n$ components of the array {\tt H\%val} may be used for
the purpose. The same applies to $\bmC$, but
there is no sensible equivalent for the non-square $\bmA$.

\subsubsection{Scaled-identity storage format}\label{scaled}
If $\bmH$ is a scalar multiple $h$ of the identity matrix, $h \bmI$,
only the value $h$ needs be stored,
and the first component of the array {\tt H\%val}
 may be used for
the purpose. The same applies to $\bmC$, but
as before there is no sensible equivalent for the non-square $\bmA$.

\subsubsection{Identity storage format}\label{identity}
If $\bmH$ is the identity matrix, $\bmI$,
no numerical data need be stored. The same applies to $\bmC$, but
once again there is no sensible equivalent for the non-square $\bmA$.

\subsubsection{Zero storage format}\label{zero}
Finally, if $\bmH$ is the zero matrix, $\bmzero$,
no numerical data need be stored. The same applies to both $\bmA$ and $\bmC$.

%%%%%%%%%%%%%%%%%%%%%%%%%%% kinds %%%%%%%%%%%%%%%%%%%%%%%%%%%

\input{kinds}

%%%%%%%%%%%%%%%%%%%%%% parallel usage %%%%%%%%%%%%%%%%%%%%%%%%

\input{parallel}

%%%%%%%%%%%%%%%%%%%%%% derived types %%%%%%%%%%%%%%%%%%%%%%%%

\galtypes
Six derived data types are accessible from the package.

%%%%%%%%%%% matrix data type %%%%%%%%%%%

\subsubsection{The derived data type for holding matrices}\label{typesmt}
The derived data type {\tt SMT\_TYPE} is used to hold the matrices $\bmH$,
$\bmA$ and $\bmC$. The components of {\tt SMT\_TYPE} used here are:

\begin{description}

\ittf{m} is a scalar component of type \integer,
that holds the number of rows in the matrix.

\ittf{n} is a scalar component of type \integer,
that holds the number of columns in the matrix.

\ittf{type} is a rank-one allocatable array of type default \character, that
is used to indicate the storage scheme used. If the dense storage scheme
(see \S\ref{dense}), is used,
the first five components of {\tt type} must contain the
string {\tt DENSE}.
For the sparse co-ordinate scheme (see \S\ref{coordinate}),
the first ten components of {\tt type} must contain the
string {\tt COORDINATE},
for the sparse row-wise storage scheme (see \S\ref{rowwise}),
the first fourteen components of {\tt type} must contain the
string {\tt SPARSE\_BY\_ROWS},
for the diagonal storage scheme (see \S\ref{diagonal}),
the first eight components of {\tt type} must contain the
string {\tt DIAGONAL}.
for the scaled-identity storage scheme (see \S\ref{scaled}),
the first fifteen components of {\tt type} must contain the
string {\tt SCALED\_IDENTITY}.
and for the identity storage scheme (see \S\ref{identity}),
the first eight components of {\tt type} must contain the
string {\tt IDENTITY}.
It is also permissible to set
the first four components of {\tt type} to the either of the strings
{\tt ZERO} or {\tt NONE}
in the case of matrix $\bmC$ to indicate that $\bmC = 0$.

For convenience, the procedure {\tt SMT\_put}
may be used to allocate sufficient space and insert the required keyword
into {\tt type}.
For example, if {\tt H} is of derived type {\tt SMT\_type}
and we wish to use the co-ordinate storage scheme, we may simply
%\vspace*{-2mm}
{\tt
\begin{verbatim}
        CALL SMT_put( H%type, 'COORDINATE', istat )
\end{verbatim}
}
%\vspace*{-4mm}
\noindent
See the documentation for the \galahad\ package {\tt SMT}
for further details on the use of {\tt SMT\_put}.

\ittf{ne} is a scalar variable of type \integer, that
holds the number of matrix entries.

\ittf{val} is a rank-one allocatable array of type \realdp\,
and dimension at least {\tt ne}, that holds the values of the entries.
Each pair of off-diagonal entries $h_{ij} = h_{ji}$ of a {\em symmetric}
matrix $\bmH$ is represented as a single entry
(see \S\ref{dense}--\ref{rowwise}); the same applies to $\bmC$.
Any duplicated entries that appear in the sparse
co-ordinate or row-wise schemes will be summed.
If the matrix is stored using the diagonal scheme (see \S\ref{diagonal}),
{\tt val} should be of length {\tt n}, and the value of the {\tt i}-th
diagonal stored in {\tt val(i)}.
If the matrix is stored using the scaled-identity scheme (see \S\ref{identity}),
{\tt val(1)} should be set to $h$.

\ittf{row} is a rank-one allocatable array of type \integer,
and dimension at least {\tt ne}, that may hold the row indices of the entries.
(see \S\ref{coordinate}).

\ittf{col} is a rank-one allocatable array of type \integer,
and dimension at least {\tt ne}, that may hold the column indices of the entries
(see \S\ref{coordinate}--\ref{rowwise}).

\ittf{ptr} is a rank-one allocatable array of type \integer,
and dimension at least {\tt m + 1}, that may hold the pointers to
the first entry in each row (see \S\ref{rowwise}).

\end{description}


%%%%%%%%%%% control type %%%%%%%%%%%

\subsubsection{The derived data type for holding control
 parameters}\label{typecontrol}
The derived data type
{\tt \packagename\_control\_type}
is used to hold controlling data. Default values may be obtained by calling
{\tt \packagename\_initialize}
(see \S\ref{subinit}),
while components may also be changed by calling
{\tt \fullpackagename\_read\-\_spec}
(see \S\ref{readspec}).
The components of
{\tt \packagename\_control\_type}
are:

\begin{description}

\itt{error} is a scalar variable of type \integer, that holds the
stream number for error messages. Printing of error messages in
{\tt \packagename\_solve} and {\tt \packagename\_terminate}
is suppressed if {\tt error} $\leq 0$.
The default is {\tt error = 6}.

\ittf{out} is a scalar variable of type \integer, that holds the
stream number for informational messages. Printing of informational messages in
{\tt \packagename\_solve} is suppressed if {\tt out} $< 0$.
The default is {\tt out = 6}.

\itt{print\_level} is a scalar variable of type \integer, that is used
to control the amount of informational output which is required. No
informational output will occur if {\tt print\_level} $\leq 0$. If
{\tt print\_level} $= 1$, a single line of output will be produced for each
iteration of the process. If {\tt print\_level} $\geq 2$, this output will be
increased to provide significant detail of each iteration.
The default is {\tt print\_level = 0}.

\itt{space\_critical} is a scalar variable of type default \logical,
that must be set \true\ if space is critical when allocating arrays
and  \false\ otherwise. The package may run faster if
{\tt space\_critical} is \false\ but at the possible expense of a larger
storage requirement. The default is {\tt space\_critical = .FALSE.}.

\itt{deallocate\_error\_fatal} is a scalar variable of type default \logical,
that must be set \true\ if the user wishes to terminate execution if
a deallocation  fails, and \false\ if an attempt to continue
will be made. The default is {\tt deallocate\_error\_fatal = .FALSE.}.

\itt{symmetric\_linear\_solver} is a scalar variable of type default \character\
and length 30, that specifies the external package to be used to
solve any general symmetric linear systems that might arise. Possible
choices are
{\tt 'sils'},
{\tt 'ma27'},
{\tt 'ma57'},
{\tt 'ma77'},
{\tt 'ma86'},
{\tt 'ma97'},
{\tt 'ssids'},
{\tt 'pardiso'},
{\tt 'wsmp'},
{\tt 'sytr'},
although only {\tt 'sils'}, {\tt 'sytr'}  and, for OMP 4.0-compliant compilers,
{\tt 'ssids'} are installed by default.
See the documentation for the \galahad\ package {\tt SLS} for further details.
The default is {\tt symmetric\_linear\_solver = 'ssids'}.

\itt{prefix} is a scalar variable of type default \character\
and length 30, that may be used to provide a user-selected
character string to preface every line of printed output.
Specifically, each line of output will be prefaced by the string
{\tt prefix(2:LEN(TRIM(prefix))-1)},
thus ignoring the first and last non-null components of the
supplied string. If the user does not want to preface lines by such
a string, they may use the default {\tt prefix = ""}.

\itt{SLS\_control} is a scalar variable argument of type
{\tt SLS\_control\_type} that is used to pass control
options to external packages used to solve any symmetric
linear systems that might arise.
See the documentation for the \galahad\ package {\tt SLS} for further details.
In particular, default values are as for {\tt SLS}.

\end{description}

%%%%%%%%%%% time type %%%%%%%%%%%

\subsubsection{The derived data type for holding timing
 information}\label{typetime}
The derived data type
{\tt \packagename\_time\_type}
is used to hold elapsed CPU and system clock times for the various parts
of the calculation. The components of
{\tt \packagename\_time\_type}
are:
\begin{description}
\itt{total} is a scalar variable of type \realdp, that gives
 the total CPU time (in seconds) spent in the package.

\itt{analyse} is a scalar variable of type \realdp, that gives
 the CPU time spent forming and analysing $\bmK$.

\itt{factorize} is a scalar variable of type \realdp, that gives
 the CPU time spent factorizing $\bmK$.

\itt{solve} is a scalar variable of type \realdp, that gives
 the CPU time spent solving the system \req{ls}.

\itt{update} is a scalar variable of type \realdp, that gives
 the CPU time spent updating the factorization.

\itt{clock\_total} is a scalar variable of type \realdp, that gives
 the total elapsed system clock time (in seconds) spent in the package.

\itt{clock\_analyse} is a scalar variable of type \realdp, that gives
 the elapsed system clock time spent forming analysing $\bmK$.

\itt{clock\_factorize} is a scalar variable of type \realdp, that gives
 the elapsed system clock time spent factorizing $\bmK$.

\itt{clock\_solve} is a scalar variable of type \realdp, that gives
 the elapsed system clock time spent solving the system \req{ls}.

\end{description}

%%%%%%%%%%% inform type %%%%%%%%%%%

\subsubsection{The derived data type for holding informational
 parameters}\label{typeinform}
The derived data type
{\tt \packagename\_inform\_type}
is used to hold parameters that give information about the progress and needs
of the algorithm. The components of
{\tt \packagename\_inform\_type}
are:

\begin{description}

\itt{status} is a scalar variable of type \integer, that gives the
exit status of the algorithm.
%See Sections~\ref{galerrors} and \ref{galinfo}
See \S\ref{galerrors}
for details.

\itt{alloc\_status} is a scalar variable of type \integer, that gives
the status of the last attempted array allocation or deallocation.
This will be 0 if {\tt status = 0}.

\itt{bad\_alloc} is a scalar variable of type default \character\
and length 80, that  gives the name of the last internal array
for which there were allocation or deallocation errors.
This will be the null string if {\tt status = 0}.

\itt{factorization\_integer} is a scalar variable of type \longinteger,
that reports the number of integers required to hold the factorization.

\itt{factorization\_real} is a scalar variable of type \longinteger,
that reports the number of reals required to hold the factorization.

\itt{rank} is a scalar variable of type \integer, that gives the
computed rank of $\bmA$.

\itt{rank\_def} is a scalar variable of type default \logical, that has the
value \true\ if {\tt \packagename\_factorize} believes that
$\bmA$ is rank defficient, and \false\ otherwise

\ittf{time} is a scalar variable of type {\tt \packagename\_time\_type}
whose components are used to hold elapsed CPU and system clock times
(in seconds) for the various parts of the calculation
(see Section~\ref{typetime}).

\itt{SLS\_inform} is a scalar variable argument of type
{\tt SLS\_inform\_type} that is used to pass information
concerning the progress of the external packages used to solve any symmetric
linear systems that might arise.
See the documentation for the \galahad\ package {\tt SLS} for further details.

\end{description}

%%%%%%%%%%% data type %%%%%%%%%%%

\subsubsection{The derived data type for holding problem data}\label{typedata}
The derived data type
{\tt \packagename\_data\_type}
are used to hold all the data for the problem and the factors of $\bmK$
between calls of
{\tt \packagename} procedures.
This data should be preserved, untouched, from the initial call to
{\tt \packagename\_initialize}
to the final call to
{\tt \packagename\_terminate}.

%%%%%%%%%%%%%%%%%%%%%% argument lists %%%%%%%%%%%%%%%%%%%%%%%%

\galarguments
There are five procedures for user calls
(see \S\ref{galfeatures} for further features):

\begin{enumerate}
\item The subroutine
      {\tt \packagename\_initialize}
      is used to set default values, and initialize private data,
      before solving one or more problems with the
      same sparsity and bound structure.
\item The subroutine
      {\tt \packagename\_analyse}
      is called to form and analyse the structure of $\bmK$ prior to 
      factorization.
\item The subroutine
      {\tt \packagename\_factorize}
      is called to factorize $\bmK$. This may be called multiple times
      for matrices with identical structure but different numerical values.
\item The subroutine
      {\tt \packagename\_solve}
      is called to apply the factorization of $\bmK$, that is to solve a 
      linear system of the form \req{ls}.
\item The subroutine
      {\tt \packagename\_terminate}
      is provided to allow the user to automatically deallocate array
       components of the private data, allocated by
       {\tt \packagename\_analyse}
       at the end of the solution process.
\end{enumerate}

%%%%%% initialization subroutine %%%%%%

\subsubsection{The initialization subroutine}\label{subinit}
 Default values are provided as follows:
\vspace*{1mm}

\hspace{8mm}
{\tt CALL \packagename\_initialize( data, control, inform )}

\vspace*{-3mm}
\begin{description}

\itt{data} is a scalar \intentinout\ argument of type
{\tt \packagename\_data\_type}
(see \S\ref{typedata}). It is used to hold data about the problem being
solved.

\itt{control} is a scalar \intentout\ argument of type
{\tt \packagename\_control\_type}
(see \S\ref{typecontrol}).
On exit, {\tt control} contains default values for the components as
described in \S\ref{typecontrol}.
These values should only be changed after calling
{\tt \packagename\_initialize}.

\itt{inform} is a scalar \intentout\ argument of type
{\tt \packagename\_inform\_type}
(see Section~\ref{typeinform}). A successful call to
{\tt \packagename\_initialize}
is indicated when the  component {\tt status} has the value 0.
For other return values of {\tt status}, see Section~\ref{galerrors}.

\end{description}

%%%%%%%%% main analysis subroutine %%%%%%

\subsubsection{The subroutine for forming and analysing the matrix $\bmK$}
The matrix $\bmK$ is formed, and its structure analysed, prior to
factorization as follows:
\vspace*{1mm}

\hspace{8mm}
{\tt CALL \packagename\_analyse( n, m, H, A, C, data, control, inform )}

%\vspace*{-3mm}
\begin{description}
\itt{n} is a scalar \intentin\ argument of type \integer\ that specifies
the number of rows of $\bmH$ (and columns of $\bmA$).

\itt{m} is a scalar \intentin\ argument of type \integer\ that specifies
the number of rows of $\bmA$ and $\bmC$.

\itt{H} is a scalar \intentin\ argument of type {\tt SMT\_type} whose
components---strictly, the value of {\tt H\%val} is not needed at this 
stage---must be set to specify the data defining the matrix $\bmH$
(see \S\ref{typesmt}).

\itt{A} is a scalar \intentin\ argument of type {\tt SMT\_type} whose
components---strictly, the value of {\tt A\%val} is not needed at this 
stage---must be set to specify the data defining the matrix $\bmA$
(see \S\ref{typesmt}).

\itt{C} is a scalar \intentin\ argument of type {\tt SMT\_type} whose
components---strictly, the value of {\tt C\%val} is not needed at this 
stage---must be set to specify the data defining the matrix $\bmC$
(see \S\ref{typesmt}).

\itt{data} is a scalar \intentinout\ argument of type
{\tt \packagename\_data\_type}
(see \S\ref{typedata}). It is used to hold data about the problem being
solved. It must not have been altered {\bf by the user} since the last call to
{\tt \packagename\_initialize}.

\itt{control} is a scalar \intentin\ argument of type
{\tt \packagename\_control\_type}
(see \S\ref{typecontrol}). Default values may be assigned by calling
{\tt \packagename\_initialize} prior to the first call to
{\tt \packagename\_solve}.

\itt{inform} is a scalar \intentout\ argument of type
{\tt \packagename\_inform\_type}
(see \S\ref{typeinform}). A successful call to
{\tt \packagename\_solve}
is indicated when the  component {\tt status} has the value 0.
For other return values of {\tt status}, see \S\ref{galerrors}.

\end{description}

%%%%%%%%% main factorization subroutine %%%%%%

\subsubsection{The subroutine for factorizing $\bmK$}
The matrix $\bmK$ is factorized as follows:
\vspace*{1mm}

\hspace{8mm}
{\tt CALL \packagename\_factorize( n, m, H, A, C, data, control, inform )}

\noindent
Components {\tt n}, {\tt m}, {\tt data} and {\tt control} 
are exactly as described for
{\tt \packagename\_analyse} and must not have been altered {\bf by the user}
in the interim.

%\vspace*{-3mm}
\begin{description}
\itt{H} is a scalar \intentin\ argument of type {\tt SMT\_type} whose
component {\tt H\%val} must be set to specify the values defining the 
matrix $\bmH$ (see \S\ref{typesmt}).

\itt{A} is a scalar \intentin\ argument of type {\tt SMT\_type} whose
component {\tt A\%val} must be set to specify the values defining the 
matrix $\bmA$ (see \S\ref{typesmt}).

\itt{C} is a scalar \intentin\ argument of type {\tt SMT\_type} whose
component {\tt C\%val} must be set to specify the values defining the 
matrix $\bmC$ (see \S\ref{typesmt}).

\itt{inform} is a scalar \intentout\ argument of type
{\tt \packagename\_inform\_type}
(see \S\ref{typeinform}). A successful call to
{\tt \packagename\_solve}
is indicated when the  component {\tt status} has the value 0.
For other return values of {\tt status}, see \S\ref{galerrors}.

\end{description}

%%%%%%%%% main solution subroutine %%%%%%

\subsubsection{The subroutine for solving linear systems involving $\bmK$}
The factorization may be applied to solve a system of the
form \req{ls} as follows:
\vspace*{1mm}

\hspace{8mm}
{\tt CALL \packagename\_solve( n, m, SOL, data, control, inform )}
\vspace*{1mm}

\noindent
Components {\tt n}, {\tt m}, {\tt data} and {\tt control} 
are exactly as described for
{\tt \packagename\_factorize} and must not have been altered {\bf by the user}
in the interim.

\vspace*{-3mm}
\begin{description}

\itt{SOL} is a rank-one  \intentinout\ array of type default \real\
and length at least {\tt n+m}, that must be set on entry to hold
the composite vector $( a^T \;\; b^T)^T$.
In particular {\tt SOL(}$i${\tt )}, $i = 1,$ \ldots {\tt n} should be
set to $a_i$, and
{\tt SOL(n}$+j${\tt )}, $j = 1, \ldots,$ {\tt m} should be
set to $b_j$. On successful exit, {\tt SOL}
will contain the solution $( x^T \;\; y^T)^T$ to \req{ls}, that is
{\tt SOL(}$i${\tt )}, $i = 1, \ldots,$ {\tt n} will give
$x_i$, and  {\tt SOL(n}$+j${\tt )}, $j = 1,\ldots,$ {\tt m} will contain
$y_j$.

\itt{inform} is a scalar \intentout\ argument of type
{\tt \packagename\_inform\_type}
(see \S\ref{typeinform}), that should be passed unaltered since
the last call to {\tt \packagename\_factorize} or
{\tt \packagename\_solve}.  A successful call to
{\tt \packagename\_solve}
is indicated when the  component {\tt status} has the value 0.
For other return values of {\tt status}, see \S\ref{galerrors}.

\end{description}

%%%%%%% termination subroutine %%%%%%

\subsubsection{The  termination subroutine}
All previously allocated arrays are deallocated as follows:
\vspace*{1mm}

\hspace{8mm}
{\tt CALL \packagename\_terminate( data, control, inform )}

%\vspace*{-3mm}
\begin{description}

\itt{data} is a scalar \intentinout\ argument of type
{\tt \packagename\_data\_type}
exactly as for
{\tt \packagename\_solve},
which must not have been altered {\bf by the user} since the last call to
{\tt \packagename\_initialize}.
On exit, array components will have been deallocated.

\itt{control} is a scalar \intentin\ argument of type
{\tt \packagename\_control\_type}
exactly as for
{\tt \packagename\_solve}.

\itt{inform} is a scalar \intentout\ argument of type
{\tt \packagename\_inform\_type}
exactly as for
{\tt \packagename\_solve}.
Only the component {\tt status} will be set on exit, and a
successful call to
{\tt \packagename\_terminate}
is indicated when this  component {\tt status} has the value 0.
For other return values of {\tt status}, see \S\ref{galerrors}.

\end{description}

%%%%%%%%%%%%%%%%%%%%%% Warning and error messages %%%%%%%%%%%%%%%%%%%%%%%%

\galerrors
A negative value of {\tt inform\%status} on exit from
{\tt \packagename\_analyse},
{\tt \packagename\_factorize},
{\tt \packagename\_solve}
or
{\tt \packagename\_terminate}
indicates that an error has occurred. No further calls should be made
until the error has been corrected. Possible values are:

\begin{description}

\itt{\galerrallocate.} An allocation error occurred.
A message indicating the offending
array is written on unit {\tt control\%error}, and the returned allocation
status and a string containing the name of the offending array
are held in {\tt inform\%alloc\_\-status}
and {\tt inform\%bad\_alloc} respectively.

\itt{\galerrdeallocate.} A deallocation error occurred.
A message indicating the offending
array is written on unit {\tt control\%error} and the returned allocation
status and a string containing the name of the offending array
are held in {\tt inform\%alloc\_\-status}
and {\tt inform\%bad\_alloc} respectively.

\itt{\galerrrestrictions.} One of the restrictions
   {\tt prob\%n} $> 0$ or {\tt prob\%m} $\geq  0$
    or requirements that {\tt prob\%A\_type}, {\tt prob\%H\_type} and
    {\tt prob\%C\_type} contain its relevant string
    {\tt 'DENSE'}, {\tt 'COORDINATE'}, {\tt 'SPARSE\_BY\_ROWS'},
    {\tt 'DIAGONAL'}, {\tt 'SCALED\_IDENTITY'}, {\tt 'IDENTITY'},
    {\tt 'ZERO'} or {\tt 'NONE'}
    has been violated.

%\itt{-3.} At least one of the arrays
% {\tt p\%A\_val}, {\tt p\%A\_row}, {\tt p\%A\_col},
% {\tt p\%H\_val}, {\tt p\%H\_row} or {\tt p\%H\_col},
% is not large enough to hold the original, or reordered, matrices $\bmA$
% or $\bmH$.

\itt{\galerranalysis.} An error was reported by {\tt SLS\_analyse}. The return
status from {\tt SLS\_analyse} is given in
{\tt inform\%SLS\_inform\%status}.
See the documentation for the \galahad\ package {\tt SLS} for further details.

\itt{\galerrfactorization.} An error was reported by {\tt SLS\_factorize}.
The return status from {\tt SLS\_factorize} is given in
{\tt inform\%SLS\_inform\%status}.
See the documentation for the \galahad\ package {\tt SLS} for further details.

\itt{\galerrsolve.} An error was reported by {\tt SLS\_solve}. The return
status from {\tt SLS\_solve} is given in {\tt inform\%SLS\_inform\%status}.
See the documentation for the \galahad\ package {\tt SLS} for further details.

\itt{\galunknownsolver.} The requested linear equation solver is not available.

\end{description}

A positive value of {\tt inform\%status} on exit from
{\tt \packagename\_factorize} warns of unexpected behaviour.
A possible values is:

\begin{description}

\itt{1.} The matrx $\bmA$ is rank defficient.

\end{description}

%%%%%%%%%%%%%%%%%%%%%% Further features %%%%%%%%%%%%%%%%%%%%%%%%

\galfeatures
\noindent In this section, we describe an alternative means of setting
control parameters, that is components of the variable {\tt control} of type
{\tt \packagename\_control\_type}
(see \S\ref{typecontrol}),
by reading an appropriate data specification file using the
subroutine {\tt \packagename\_read\_specfile}. This facility
is useful as it allows a user to change  {\tt \packagename} control parameters
without editing and recompiling programs that call {\tt \packagename}.

A specification file, or specfile, is a data file containing a number of
"specification commands". Each command occurs on a separate line,
and comprises a "keyword",
which is a string (in a close-to-natural language) used to identify a
control parameter, and
an (optional) "value", which defines the value to be assigned to the given
control parameter. All keywords and values are case insensitive,
keywords may be preceded by one or more blanks but
values must not contain blanks, and
each value must be separated from its keyword by at least one blank.
Values must not contain more than 30 characters, and
each line of the specfile is limited to 80 characters,
including the blanks separating keyword and value.



The portion of the specification file used by
{\tt \packagename\_read\_specfile}
must start
with a "{\tt BEGIN \packagename}" command and end with an
"{\tt END}" command.  The syntax of the specfile is thus defined as follows:
\begin{verbatim}
  ( .. lines ignored by SSLS_read_specfile .. )
    BEGIN SSLS
       keyword    value
       .......    .....
       keyword    value
    END
  ( .. lines ignored by SSLS_read_specfile .. )
\end{verbatim}
where keyword and value are two strings separated by (at least) one blank.
The ``{\tt BEGIN \packagename}'' and ``{\tt END}'' delimiter command lines
may contain additional (trailing) strings so long as such strings are
separated by one or more blanks, so that lines such as
\begin{verbatim}
    BEGIN SSLS SPECIFICATION
\end{verbatim}
and
\begin{verbatim}
    END SSLS SPECIFICATION
\end{verbatim}
are acceptable. Furthermore,
between the
``{\tt BEGIN \packagename}'' and ``{\tt END}'' delimiters,
specification commands may occur in any order.  Blank lines and
lines whose first non-blank character is {\tt !} or {\tt *} are ignored.
The content
of a line after a {\tt !} or {\tt *} character is also
ignored (as is the {\tt !} or {\tt *}
character itself). This provides an easy manner to "comment out" some
specification commands, or to comment specific values
of certain control parameters.

The value of a control parameters may be of three different types, namely
integer, logical or real.
Integer and real values may be expressed in any relevant Fortran integer and
floating-point formats (respectively). Permitted values for logical
parameters are "{\tt ON}", "{\tt TRUE}", "{\tt .TRUE.}", "{\tt T}",
"{\tt YES}", "{\tt Y}", or "{\tt OFF}", "{\tt NO}",
"{\tt N}", "{\tt FALSE}", "{\tt .FALSE.}" and "{\tt F}".
Empty values are also allowed for
logical control parameters, and are interpreted as "{\tt TRUE}".

The specification file must be open for
input when {\tt \packagename\_read\_specfile}
is called, and the associated device number
passed to the routine in device (see below).
Note that the corresponding
file is {\tt REWIND}ed, which makes it possible to combine the specifications
for more than one program/routine.  For the same reason, the file is not
closed by {\tt \packagename\_read\_specfile}.

Control parameters corresponding to the components
{\tt SLS\_control}
may be changed by including additional sections enclosed by
``{\tt BEGIN SLS}'' and
``{\tt END SLS}''.
See the specification sheets for the package
{\tt \libraryname\_SLS}
for further details.

\subsubsection{To read control parameters from a specification file}
\label{readspec}

Control parameters may be read from a file as follows:
\hskip0.5in

\def\baselinestretch{0.8}
{\tt
\begin{verbatim}
     CALL SSLS_read_specfile( control, device )
\end{verbatim}
}
\def\baselinestretch{1.0}

\begin{description}
\itt{control} is a scalar \intentinout argument of type
{\tt \packagename\_control\_type}
(see \S\ref{typecontrol}).
Default values should have already been set, perhaps by calling
{\tt \packagename\_initialize}.
On exit, individual components of {\tt control} may have been changed
according to the commands found in the specfile. Specfile commands and
the component (see \S\ref{typecontrol}) of {\tt control}
that each affects are given in Table~\ref{specfile}.
\bctable{|l|l|l|}
\hline
  command & component of {\tt control} & value type \\
\hline

  {\tt error-printout-device} & {\tt \%error} & integer \\
  {\tt printout-device} & {\tt \%out} & integer \\
  {\tt print-level} & {\tt \%print\_level} & integer \\
  {\tt space-critical}   & {\tt \%space\_critical} & logical \\
  {\tt deallocate-error-fatal}   & {\tt \%deallocate\_error\_fatal} & logical \\
  {\tt symmetric-linear-equation-solver} & {\tt \%symmetric\_linear\_solver} & character \\
  {\tt output-line-prefix} & {\tt \%prefix} & character \\
\hline

\ectable{\label{specfile}Specfile commands and associated
components of {\tt control}.}

%Individual components of the components {\tt control\%SLS\_control}
%and {\tt control\%ULS\_control}
%may be changed by specifying separate specfiles of the form
%\begin{verbatim}
%    BEGIN SLS
%     .......
%    END
%\end{verbatim}
%and
%\begin{verbatim}
%    BEGIN ULS
%     .......
%%    END
%\end{verbatim}
%just as described in the documentation for the \galahad\ packages
%{\tt SLS} and {\tt ULS}.

\itt{device} is a scalar \intentin argument of type \integer,
that must be set to the unit number on which the specfile
has been opened. If {\tt device} is not open, {\tt control} will
not be altered and execution will continue, but an error message
will be printed on unit {\tt control\%error}.

\end{description}

%%%%%%%%%%%%%%%%%%%%%% Information printed %%%%%%%%%%%%%%%%%%%%%%%%

\galinfo
If {\tt control\%print\_level} is positive, information about the progress
of the algorithm will be printed on unit {\tt control\-\%out}.
If {\tt control\%print\_level} $= 1$, statistics concerning the factorization,
as well as warning and error messages will be reported.
If {\tt control\%print\_level} $= 2$, additional information about the
progress of the factorization and the solution phases will be given.
If {\tt control\%print\_level} $> 2$, debug information, of little
interest to the general user, may be returned.

%%%%%%%%%%%%%%%%%%%%%% GENERAL INFORMATION %%%%%%%%%%%%%%%%%%%%%%%%

\galgeneral

\galcommon None.
\galworkspace Provided automatically by the module.
\galroutines None.
\galmodules {\tt \packagename\_solve} calls the \galahad\ packages
{\tt GALAHAD\_CLOCK},
{\tt GALAHAD\_SY\-M\-BOLS}, \\
{\tt GALAHAD\-\_SPACE},
{\tt GALAHAD\_SMT},
{\tt GALAHAD\_QPT},
{\tt GALAHAD\_SLS},
{\tt GALAHAD\_ULS} and
{\tt GALAHAD\_SPECFILE},
\galio Output is under control of the arguments
 {\tt control\%error}, {\tt control\%out} and {\tt control\%print\_level}.
\galrestrictions {\tt prob\%n} $> 0$, {\tt prob\%m} $\geq  0$,
{\tt prob\%H\_type}, {\tt prob\%A\_type}
and {\tt prob\%C\_type} $\in \{${\tt 'DENSE'},
 {\tt 'COORDINATE'}, {\tt 'SPARSE\_BY\_ROWS'}, {\tt 'DIAGONAL'},
 {\tt 'SCALED\_IDENTITY'}, {\tt 'IDENTITY'} $\}$,
 {\tt 'ZERO'}, {\tt 'NONE'}.
\galportability ISO Fortran~95 + TR 15581 or Fortran~2003.
The package is thread-safe.

%%%%%%%%%%%%%%%%%%%%%% METHOD %%%%%%%%%%%%%%%%%%%%%%%%

\galmethod
The method simply assembles $\bmK$ from its components, and then relies
on {\tt SLS} for analysis, factorization and solves.

%%%%%%%%%%%%%%%%%%%%%% EXAMPLE %%%%%%%%%%%%%%%%%%%%%%%%

\galexample
Suppose we wish to solve the linear system \req{ls} with matrix data
\disp{\bmH = \mat{ccc}{1 & & 4 \\ & 2 & \\ 4 &  & 3}, \;\;
 \bmA = \mat{ccc}{ 2 & 1 & \\ & 1 & 1}
\tim{and}  \bmC = \mat{cc}{  & 1 \\ 1 &   }}
and right-hand sides
\disp{\bma = \vect{7 \\ 4 \\ 8} \tim{and} \bmb = \vect{2 \\ 1}.}
Then storing the matrices in sparse co-ordinate format,
we may use the following code:
\renewcommand{\packagedir}{\galahaddir/src/forthcoming/\package}
{\tt \small
\VerbatimInput{\packageexample}
}
\noindent
This produces the following output:
{\tt \small
\VerbatimInput{\packageresults}
}
\noindent
The same problem may be solved holding the data in
a sparse row-wise storage format by replacing the lines
{\tt \small
\begin{verbatim}
!  sparse co-ordinate storage format
...
! problem data complete
\end{verbatim}
}
\noindent
by
{\tt \small
\begin{verbatim}
! sparse row-wise storage format
   CALL SMT_put( H%type, 'SPARSE_BY_ROWS', s )  ! Specify sparse-by-rows
   CALL SMT_put( A%type, 'SPARSE_BY_ROWS', s )  ! storage for H, A and C
   CALL SMT_put( C%type, 'SPARSE_BY_ROWS', s )
   ALLOCATE( H%val( h_ne ), H%col( h_ne ), H%ptr( n + 1 ) )
   ALLOCATE( A%val( a_ne ), A%col( a_ne ), A%ptr( m + 1 ) )
   ALLOCATE( C%val( c_ne ), C%col( c_ne ), C%ptr( m + 1 ) )
   H%val = (/ 1.0_wp, 2.0_wp, 3.0_wp, 4.0_wp /) ! matrix H
   H%col = (/ 1, 2, 3, 1 /)                     ! NB lower triangular
   H%ptr = (/ 1, 2, 3, 5 /)                     ! Set row pointers
   A%val = (/ 2.0_wp, 1.0_wp, 1.0_wp, 1.0_wp /) ! matrix A
   A%col = (/ 1, 2, 2, 3 /)
   A%ptr = (/ 1, 3, 5 /)                        ! Set row pointers
   C%val = (/ 1.0_wp /)                         ! matrix C
   C%col = (/ 1 /)                              ! NB lower triangular
   C%ptr = (/ 1, 1, 2 /)                        ! Set row pointers
! problem data complete
\end{verbatim}
}
\noindent
or using a dense storage format with the replacement lines
{\tt \small
\begin{verbatim}
! dense storage format
   CALL SMT_put( H%type, 'DENSE', s )  ! Specify dense
   CALL SMT_put( A%type, 'DENSE', s )  ! storage for H, A and C
   CALL SMT_put( C%type, 'DENSE', s )
   ALLOCATE( H%val( n * ( n + 1 ) / 2 ) )
   ALLOCATE( A%val( n * m ) )
   ALLOCATE( C%val( m * ( m + 1 ) / 2 ) )
   H%val = (/ 1.0_wp, 0.0_wp, 2.0_wp, 4.0_wp, 0.0_wp, 3.0_wp /) ! H
   A%val = (/ 2.0_wp, 1.0_wp, 0.0_wp, 0.0_wp, 1.0_wp, 1.0_wp /) ! A
   C%val = (/ 0.0_wp, 1.0_wp, 0.0_wp /)                         ! C
! problem data complete
\end{verbatim}
}
\noindent
respectively.

If instead $\bmH$ had been the diagonal matrix
\disp{\bmH = \mat{ccc}{1 & &   \\ & 0 & \\  &  & 3}}
but the other data is as before, the diagonal storage scheme
might be used for $\bmH$, and in this case we would instead
{\tt \small
\begin{verbatim}
   CALL SMT_put( prob%H%type, 'DIAGONAL', s )  ! Specify dense storage for H
   ALLOCATE( p%H%val( n ) )
   p%H%val = (/ 1.0_wp, 0.0_wp, 3.0_wp /) ! Hessian values
\end{verbatim}
}
\noindent
Notice here that zero diagonal entries are stored.

\end{document}
